DANCE: An open-source analysis pipeline and low-cost hardware to quantify aggression and courtship in Drosophila

Version 1

Abstract

Abstract

Quantifying animal behaviors is pivotal for identifying the underlying neuronal and genetic mechanisms. Computational approaches have enabled automated analysis of complex behaviors such as aggression and courtship in Drosophila. However, existing approaches rely on rigid, rule-based algorithms and expensive hardware, limiting sensitivity to behavioral variations and accessibility. Here, we describe the DANCE (Drosophila Aggression and Courtship Evaluator), a low-cost, open-source platform combining machine learning-based classifiers and inexpensive hardware to quantify aggression and courtship. DANCE consists of six novel behavioral classifiers trained using a supervised machine learning algorithm. DANCE classifiers outperform existing rule-based algorithms by capturing dynamic behavioral variations. DANCE hardware is constructed using repurposed medicine blister packs and acrylic sheets, with recordings performed using smartphones, making it affordable and accessible. Benchmarking demonstrated that DANCE hardware performs comparably to sophisticated, high-cost setups. We validated DANCE in diverse contexts, including social isolation versus enrichment that modulate aggression and courtship, RNAi-mediated down-regulation of neuropeptide Dsk, and optogenetic silencing of dopaminergic neurons which promoted aggression. DANCE provides a cost-effective and portable solution for studying Drosophila behaviors in resource-limited settings or closer to natural habitats. Its accessibility and robust performance democratizes behavioral neuroscience, enabling rapid screening of genes and neuronal circuits underlying complex social behaviors.

Introduction

Detailed and accurate annotation and analysis of complex behaviors is necessary for understanding the underlying neural and molecular mechanisms. Fruit fly Drosophila melanogaster is one of the most accessible and well-studied model organisms for identifying neuronal and molecular underpinnings of behavior. Multiple large-scale screens have been conducted using Drosophila, to study complex social behaviors such as aggression and courtship (Asahina, 2017; Greenspan and Ferveur, 2000; Hall, 2002; Kravitz and Fernandez, 2015) to identify underlying neural circuitry (Agrawal et al., 2020; Asahina et al., 2014; Davis et al., 2018; Hoopfer et al., 2015; Yadav et al., 2024) and genes involved (Agrawal et al., 2020; Benzer, 1967; Gill, 1963; Hall, 1978; Ishii et al., 2022; Wang et al., 2008). These behaviors exhibit stereotyped patterns. For instance, aggression involves chasing, fencing (Jacobs, 1960), wing threats, boxing (Dow and Schilcher, 1975), lunging, and tussling (Hoffmann, 1987a, 1987b). Similarly, courtship consists of multiple stereotyped behaviors exhibited by the male fly, such as orienting, circling, and following the female (Cook and Cook, 1975; Markow, 1987; O’Dell, 2003). To stimulate the female to be more receptive, the male produces a species-specific song by vibrating and extending its wing (Bennet-Clark and Ewing, 1969; Swain and von Philipsborn, 2021). The male then attempts copulation by curling its abdomen and finally mounts the female for copulation (Bastock and Manning, 1955; Spieth, 1974).

Manual analysis by trained observers is considered the gold standard in behavioral analysis, but it is a time-consuming process unsuitable for large-scale screens (Gomez- Marin et al., 2014; Robie et al., 2017a). ‘Computational ethology’ (Anderson and Perona, 2014; Datta et al., 2019) helps address this challenge by automating behavioral annotation by leveraging advances in computer vision and machine learning (Robie et al., 2017b). This enables high-throughput behavioral screening to identify responsible genes and circuits. A typical computational ethology workflow involves recording animal behaviors and tracking their positions along with body movements. This is followed by the analysis and classification of the observed behaviors from hundreds to thousands of video frames capturing behavioral instances. Several software such as Ctrax, Caltech FlyTracker, Deep Lab Cut etc. (Branson et al., 2009; Eyjolfsdottir et al., 2014; Mathis et al., 2018) are used widely for tracking behaviors in Drosophila. Each comes with their strengths and weaknesses.

For instance, Ctrax (Branson et al., 2009) helps track fly body position and movements but leads to frequent identity switches among tracked flies, which is not the case while using FlyTracker (Eyjolfsdottir et al., 2014). Effectiveness of various machine learning pipelines is eventually measured by comparing their output to human annotation, called— ‘ground- truthing’. Rule-based algorithms such as CADABRA (Dankert et al., 2009) is used to quantify aggression, but it can lead to mis-scoring and identity switches, as revealed by ground-truthing (Simon and Heberlein, 2020), which needs to be corrected in a semi- automated manner (Kim et al., 2018). MateBook (Ribeiro et al., 2018) is another rule-based algorithm to quantify courtship; however, similar to CADABRA, it tends to miss true-positive events, leading to significant mis-scoring of behaviors.

Janelia Automatic Animal Behavior Annotator (JAABA) (Kabra et al., 2013) addresses the challenges of rigid rule-based approaches by employing a supervised learning approach. In the JAABA pipeline, user-labeled data is utilized for training to encompass the dynamic variations in behaviors, allowing it to predict behaviors based on learning from input data.

Recently, a few JAABA-based behavioral classifiers have been developed for measuring aggression (Chowdhury et al., 2021; Leng et al., 2020) and courtship (Gil-Martí et al., 2023; Pantalia et al., 2023). However, these studies either do not make these classifiers publicly available (Gil-Martí et al., 2023; Leng et al., 2020; Pantalia et al., 2023) or require specialized hardware such as 3D printed parts (Chowdhury et al., 2021; Gil-Martí et al., 2023) and expensive machine vision cameras (Chowdhury et al., 2021; Leng et al., 2020), making them inaccessible for resource-limited settings and limiting their wider adoption.

Here we describe DANCE (Drosophila Aggression and Courtship Evaluator), an open-source, user-friendly analysis and hardware pipeline to simplify and automate the process of robustly quantifying aggression and courtship behaviors. DANCE has two components: 1. A set of robust, machine vision-based behavioral classifiers developed using JAABA to quantify aggression and courtship; and 2. An inexpensive hardware utilizing off-the-shelf components to construct behavioral arenas and smartphone cameras to record behaviors. DANCE classifiers outperform previous methods (Dankert et al., 2009; Ribeiro et al., 2018), improving accuracy and reliability, and DANCE hardware eliminates the need for complex, specialized behavioral arenas and expensive machine vision cameras to assess fly behaviors. Therefore, the DANCE assay can be useful for wider adoption, including in resource-strapped settings. DANCE provides a platform for rapid behavioral screening using the Drosophila model for discovering mechanisms underlying complex social behaviors and neurological disorders.

Results

DANCE assay analysis pipeline

To overcome the challenge of time-consuming manual behavioral annotation or resource- intensive, complex hardware, we developed an automated, high-throughput quantification pipeline—DANCE and trained new behavioral classifiers using an existing machine learning algorithm- JAABA (Kabra et al., 2013), to robustly quantify aggression and courtship in Drosophila (Figure 1). We built a novel hardware platform consisting of repurposed transparent medicine tablet foils/blister packs, acrylic sheets, and paper tape to easily record aggression and courtship behaviors. To record these behaviors we used Android smartphone cameras and an electronic tablet or smartphone served as a backlight illumination source (Figure 1A, Materials and Methods). We quantify these behaviors using our ‘DANCE classifiers’. Unlike existing set-ups that cost approximately USD 3500, ‘DANCE hardware’ is fabricated using off-the-shelf components and costs less than USD 30 cents (Supplementary file 1). To benchmark the performance of the DANCE classifiers, we used pre-existing setups and rule-based methods for quantifying courtship and aggression and compared their performance with DANCE classifiers (Figure 1B-C).

DANCE assay overcomes the limitation of existing methods for quantification of aggression and courtship behaviors.

(A) Existing hardware for acquisition of aggression and courtship behavior using machine vision cameras and the simplified DANCE hardware. (B) Steps for developing DANCE classifiers and benchmarking with existing methods and manual ground-truthing to produce behavioral scores. (C) Various behavioral classifiers developed to quantify male aggression (lunge) and male courtship (wing extension, circling, following, attempted copulation, and copulation). (D) Raster plots comparing performance of ground-truth vs. DANCE classifier vs. CADABRA vs. Divider assay for aggression in a representative video. (E) Raster plots for various courtship behaviors comparing performance of ground-truth vs. DANCE classifiers vs. MateBook from representative videos. Created in BioRender.

To train DANCE classifiers using JAABA (Kabra et al., 2013), for aggressive lunges, we used a pre-existing setup described in (Dankert et al., 2009) modified from (Dierick, 2007) (Figure 1A; Figure 1— figure supplement 1); and for courtship behaviors, we used a pre- existing set up described in (Koemans et al., 2017) (Figure 1A; Figure 1— figure supplement 2). We tracked position, motion, and interactions of pairs of flies across video frames using the Caltech FlyTracker (Eyjolfsdottir et al., 2014). To avoid data leak, we randomly divided acquired videos into two categories: "training videos" and “test videos” to train and evaluate DANCE classifiers. These test videos were also manually ‘ground- truthed’ frame by frame, which is considered the gold standard for behavioral annotation (Figure 1B, D-E). We benchmarked the performance of DANCE classifiers using existing rule-based algorithms, CADABRA (Dankert et al., 2009) and MateBook (Ribeiro et al., 2018); and an existing JAABA aggression classifier (Chowdhury et al., 2021). We then compared their output against manual ground-truth. Figure 1 D-E show examples of courtship and aggression video annotation, suggesting that DANCE classifiers outperform existing methods and are comparable with manual ground-truth. Subsequent sections describe the quantitative analysis of individual DANCE classifiers and benchmarking of DANCE hardware.

DANCE lunge classifier to quantify aggressive behavior

Aggression is an innate, complex behavior, and Drosophila males exhibit several stereotyped behavioral patterns during aggressive encounters, with lunging used widely as a measure of overall aggression in males (Agrawal et al., 2020; Asahina et al., 2014; Chiu et al., 2021; Chowdhury et al., 2021; Davis et al., 2014; Dierick, 2007; Hoffmann, 1987a; Hoopfer et al., 2015; Hoyer et al., 2008; Jung et al., 2020; Nilsen et al., 2004; Watanabe et al., 2017; Yadav et al., 2024). A lunge is defined as a male fly raising its front legs and hitting down on the other fly. We developed a new classifier using JAABA (Kabra et al., 2013) to robustly quantify aggressive lunges in Drosophila, hereafter referred to as the ‘DANCE lunge classifier.’ We quantified lunges using our classifier from 20-minute-long videos and compared the output with manual ground-truthing and existing methods— CADABRA (Dankert et al., 2009) and ‘Divider assay classifier’ (Chowdhury et al., 2021). CADABRA tends to miss several true positive lunges, likely due to its rigid, rule-based framework, which cannot adapt to the dynamic variations in behavior. To address this, ‘Divider assay classifier’ was developed based on JAABA, but the setup requires use of 3- D printed rectangular arena and dividers as well as expensive machine vision cameras (Chowdhury et al., 2021).

Figure 2A shows lunge scores from 40 different videos using ground-truth, DANCE lunge classifier, CADABRA, and Divider assay classifier. While the ground-truth and DANCE classifier’s output are comparable, CADABRA and the Divider assay classifier underscored lunges across videos. We ground truthed the DANCE lunge classifier against 40 ‘test videos’ (Figure 2A, Materials and Methods). To address potential observer bias during ground-truthing, we compared the annotations between two separate evaluators for the DANCE lunge classifier and found no significant differences (Figure 2— figure supplement 1A). Since aggressive lunges have a large dynamic range, we benchmarked our classifier across a range of aggressive behaviors, and subdivided the ground truthed videos into four categories: 1. low aggressive, 0–70 lunges (Figure 2B); 2. moderately aggressive, 71–160 lunges (Figure 2C); 3. highly aggressive, 161–300 lunges (Figure 2D); and 4. hyper- aggressive, >300 lunges (Figure 2E). We found that DANCE lunge classifier and manual ground-truth performance was comparable across all four categories (Figure 2 B-E). In contrast, the CADABRA and Divider assay classifier showed significant underscoring from the manual ground-truth (Figure 2B-E), however, a strong correlation was observed between our DANCE lunge classifier and ground-truth (Figure 2F, Supplementary file 2), whereas, CADABRA and the Divider assay classifier showed only a weaker correlation with the ground truth (Figure 2G-H, Supplementary file 2). We reasoned that CADABRA’s lower performance is most likely due to the rigid rules used to define a lunge (Dankert et al., 2009) (Materials and Methods); and the Divider assay classifier’s trained in a rectangular arena did not perform as well in a circular arena since JAABA classifiers utilize features related to arena geometry.

DANCE lunge classifier outperforms existing methods of quantifying male aggression.

(A) Lunge scores from 20 minute long videos from ground-truth (grey), DANCE lunge classifier (orange), CADABRA (purple) and Divider assay classifier (green). (B-E) Comparison of lunge scores from different levels of aggression. Male flies showing lunges ranging from 0 to 500 were compared manually, with a new DANCE lunge classifier, CADABRA, and Divider assay classifier. (B) ‘low aggressive’, (n=10; **P<0.0017, ****P<0.0001). (C) 71–160 lunges, ’moderately aggressive’, (n=11; **P<0.0102, ***P<0.0002, ***P<0.0001). (D) 161–300 lunges, ‘highly aggressive’, (n = 11; **P<0.0057, ***P<0.0002, ***P<0.0004). and (E) >300 lunges, ‘hyper-aggressive’, (n = 8; *P<0.0402, **P<0.0029, ***P<0.0006; Friedman’s ANOVA with Dunn’s test) (F) Regression analysis of DANCE ‘lunge classifier’ vs. manual scores (R2=0.9893, n=40). (G) CADABRA vs. DANCE lunge classifier (R2=0.9, n=40). (H) Divider assay lunge classifier vs. manual scores (R2=0.7739, n=40). (I) F1 score, precision, and recall between DANCE lunge classifier, CADABRA, and Divider assay classifier.

To further evaluate the performance of our DANCE lunge classifier, we used three metrics— precision, recall, and F1 score and compared the output of the classifier and manual ground-truth across 40 test videos. We calculated the counts of true positives, false positives, and false negatives to derive precision, which depicts the ratio of true positives among overall predictions from the classifier; recall, which depicts the ratio of true positives among the ground-truth scores; and F1 score, which is a harmonic mean of precision and recall (Figure 2I) (Materials and Methods). Our DANCE lunge classifier had a precision of 76.80%, a recall rate of 71.53% and an overall F1 score of 73.60% (Figure 2I), which is higher than the CADABRA and the Divider assay suggesting that the DANCE lunge classifier outperforms these existing methods. Together, our analysis suggests that the DANCE lunge classifier performs with high precision and quantifies lunge numbers robustly over a broad range of fighting intensities.

DANCE classifiers to quantify courtship behaviors in Drosophila

The first report of Drosophila courtship behavior, identified stereotypic behaviors such as wing ‘scissor-like’ movements, males ‘swaying around the female,’ licking, tapping, and mounting (Sturtevant, 1915). By the 2000s, studies uncovered genes and neural circuits involved in courtship (Dickson, 2008; Pavlou and Goodwin, 2013). Automated analysis techniques for courtship exist but face barriers to adoption due to expensive hardware, reliance on 3D printing, or lack of publicly available analysis code or classifiers (Gil-Martí et al., 2023; Koemans et al., 2017; Leng et al., 2020; Reza et al., 2013).

MateBook is a recent rule-based algorithm that was developed to automate quantifying various courtship behaviors in Drosophila males (Ribeiro et al., 2018). It relies on predefined mathematical rules based on CADABRA (Dankert et al., 2009). To resolve ambiguities in the two overlapping flies during estimating their trajectories from the video recordings, it assigns the identities by relying on the distinct body sizes of the male and female fly, as the females are larger than males. However, decapitated virgin females are of similar body size as males, and are often used in courtship assays to assess the male attractiveness to other factors such as pheromones, independent of the female behavior (Cook and Cook, 1975; Spieth, 1966). In such cases, rigid rule-based approach can introduce both false positives and false negatives since the condition about differential body size is not met.

To address the inherent inflexibility of rule-based algorithms, we trained and validated five new courtship classifiers using JAABA (Kabra et al., 2013) to provide a detailed and robust analysis of multiple steps of courtship ritual (Sokolowski, 2001) including— wing extension, following, circling, attempted copulation, and copulation. We evaluated the performance of our DANCE classifiers by comparing the output to ground-truth and MateBook (Ribeiro et al., 2018). Criteria used to develop various classifiers are described in Materials and Methods. To compare the output, we calculated a behavior index, since courtship behaviors, unlike aggressive lunges, are of varied duration. To address potential observer bias during ground-truthing, we compared the annotations between two separate evaluators for the DANCE courtship classifiers and found no significant differences, suggesting robustness of our classifiers (Figure 2— figure supplement 1B-F).

Wing Extension

During unilateral wing extension, the male vibrates its wing at a specific frequency to produce a species-specific courtship song to attract the female (Shorey, 1962; Spieth, 1952). Figure 3A shows wing extension index from the ground-truth (grey) and the ‘DANCE wing extension classifier’ (orange), which are comparable across 15 videos from males courting decapitated virgin females, however, MateBook’s output (purple) seems to underscore these events in most videos (Figure 3A, Friedman’s ANOVA with Dunn’s test; ground-truth vs. MateBook, p=0.0020, n=15; ground-truth vs. DANCE classifier, p>0.9999) (Figure 3A). We also evaluated the overall wing extension index from all 15 videos and found ground-truth scores and the DANCE wing extension classifier matched well, but MateBook significantly underscored wing extension (Figure 3B). We found a strong correlation between the DANCE wing extension classifier index to ground-truth (Figure 3C) but a weaker correlation with MateBook (Figure 3D). To further evaluate our classifier’s performance, we calculated precision, recall, and F1 scores, as described earlier. The DANCE wing extension classifier had a precision of 92.19%, a recall rate of 98.09%, and an overall F1 score of 95.05% when compared to the ground-truth which is higher than the MateBook (Figure 3E).

DANCE wing extension classifier outperforms existing methods of quantification.

(A) Wing extension index of males from 15 minutes long videos from the ground-truth (grey), DANCE wing extension classifier (orange) and MateBook (purple), against decapitated virgin females. MateBook underscored wing extension across multiple videos (black arrows). (B) Comparison of ground-truth vs. DANCE vs. MateBook wing extension classifier (Kruskal-Wallis ANOVA with Dunn’s test, ns, p>0.9999, *p=0.0436; n=15). (C) Regression analysis of the DANCE wing extension classifier vs. ground-truth (R2=0.9831, n=15). (D) MateBook vs. ground-truth (R2=0.1054, n=15). (E) F1 score, precision, and recall of DANCE wing extension classifier and MateBook against ground- truth scores.

Quantification of the videos containing mated females identified a robust correlation between our classifier’s output and the ground-truth, whereas a general trend of underscoring was observed for MateBook (Figure 3— figure supplement 1). Together, this suggests that the DANCE wing extension classifier is robust for both decapitated and mated females and performs with high precision.

Attempted copulation and copulation

Copulation in male flies lasts for about 15-25 minutes, and its duration is primarily determined by the male (MacBean and Parsons, 1967). Interrupted mating experiments show that the sperms are transferred several minutes after copulation begins (Fowler, 1973; Tompkins et al., 1980). Thus this differentiates the mounting behavior into two— successful copulation and unsuccessful attempted-copulation. We developed the ‘DANCE attempted copulation’ with a threshold of 0.33 to <45 seconds and the ‘DANCE copulation’ classifiers with a threshold of >45 seconds, based on existing definitions (Ribeiro et al., 2018), but with modifications, since MateBook only quantifies ‘copulation’ and not ‘attempted-copulation’ (Materials and Methods). For ground-truthing, we combined both the dataset of mated and decapitated virgin females, because we observed that MateBook consistently overscored in both datasets. (Figure 4A-B). MateBook’s rule-based approach led to significant overscoring with reduced thresholds for attempted copulations, especially with decapitated virgins, due to both increased false positives and identity switches. As expected, the DANCE classifier matched well with the ground-truth scores for each individual video (Figure 4A— Friedman’s ANOVA with Dunn’s test; p>0.9999, n=32), as well as the overall attempted copulation index (Figure 4B). We found a strong correlation between ground-truth and the DANCE attempted copulation classifier index (Figure 4C) but a weaker correlation with MateBook (Figure 4D). The DANCE classifier performed robustly with a precision rate of 82.55%, a recall rate of 89.24%, and a F1 score of 85.77% as compared to MateBook (Figure 4E). Together, this suggests that the DANCE attempted copulation outperforms existing rule based framework and is able to capture dynamic variations in behaviors robustly.

DANCE attempted-copulation classifier outperforms existing methods of quantification.

(A) Attempted copulation index of males from 15 minute long videos from the ground-truth (grey), ‘DANCE attempted copulation classifier’ (orange) and MateBook (purple) against both mated and decapitated females. (B) Comparison of ground-truth vs. DANCE attempted-copulation classifier vs. MateBook (Kruskal-Wallis ANOVA with Dunn’s test, ns, p>0.9999; ****p<0.0001, n=32). (C) Regression analysis of the attempted-copulation classifier vs. ground-truth (R2=0.8742, n=32). (D) Regression analysis of MateBook vs. ground-truth (R2=1512, n=32). (E) F1 score, precision, and recall of DANCE and MateBook attempted-copulation classifiers against ground-truth scores.

Since the male readily copulates with the virgin females for several minutes, we used videos where males were paired with virgin females to train the ‘DANCE copulation classifier’. Our classifier matched well with the ground truth in the number of bouts, with a 100% precision, recall and F1 score (Figure 4— figure supplement 1). The rule-based approach of MateBook also works well with copulation, which MateBook defines as ‘occlusion’ persisting for a long time (Materials and Methods). However sometimes, misaligned videos can cause ‘arena detection’ errors, causing missed events (Figure 4— figure supplement 1A, video#18) or false positives (Figure 4— figure supplement 1, video#13) leading to lower precision, recall and F1 scores.

Circling

The male circles around the female, which is a unique male behavior during courtship. Studies have shown that a refractory female gets restimulated by the circling behavior of the males (Kessler, 1962). In addition, differences in the frequency of circling behavior and other courtship rituals is used to maintain reproductive isolation among different species (Brown, 1965).

To evaluate the ‘DANCE circling classifier’, we used both the decapitated virgin female (Figure 5) and mated female dataset (Figure 5— Figure Supplement 1). The circling index from the ground-truth and the DANCE classifier output are comparable; however, MateBook underscored these events (Figure 5A), most likely due to identity switches. DANCE circling classifier matched well from the ground-truth for individual videos (Figure 5A, Friedman’s ANOVA with Dunn’s test; p<0.2049, n=12). Overall circling index from all 12 videos from the ground-truth scores and the DANCE circling classifier were comparable (Figure 5B), whereas MateBook significantly underscores circling in individual videos (Figure 5A, Friedman’s ANOVA with Dunn’s test; ****p<0.0001, n=12). We found a strong correlation between the DANCE circling classifier index to ground-truth (Figure 5C) but a weaker correlation with MateBook (Figure 5D). The DANCE circling classifier has a precision rate and a recall rate of 98.04% and 92.09%, with a F1 score of 94.97% (Figure 5E), suggesting robustness of the classifier.

DANCE circling classifier outperforms existing methods of quantification.

(A) Circling index of males from 15 minutes long videos from the ground-truth (grey), ‘DANCE circling classifier’ (orange) and MateBook (purple) against decapitated virgin females. (B) Comparison of manual vs. DANCE vs. MateBook circling classifier (Ordinary one-way ANOVA with Dunnett’s test, ns, p=0.8014; *p=0.0157, n=12). (C) Regression analysis of the circling classifier vs. ground-truth (R2=0.92, n=12). (D) MateBook vs. ground-truth (R2=0.88, n=12). (E) F1 score, precision, and recall of DANCE and MateBook circling classifiers against ground-truth scores.

Following

During ‘following’, the male follows around the female if it is moving in order to get closer to it and initiates the other courtship behaviors such as wing extension, licking, etc. (Spieth, 1968). Since following is a relatively simpler behavior, both MateBook and the ‘DANCE following classifier’ performed well when compared to ground-truth (Figure 5— figure supplement 2A-D). However, the DANCE classifier had better precision, recall, and F1 scores (91.16%, 91.07%, and 91.12%) as compared to MateBook (65.77%, 83.39%, and 73.53%), indicating lower number of false positives and false negatives (Figure 5— figure supplement 2E).

DANCE hardware

To measure aggression and courtship in flies, existing setups (Figure 1A, Figure 1— figure supplement 1-2) (Dankert et al., 2009; Koemans et al., 2017) have several bottlenecks, making it challenging to adopt these assays by the broader neuroscience community. For instance, it requires complex and specialized hardware (Dankert et al., 2009; Koemans et al., 2017), necessitating access to sophisticated machine shops; or 3D printers to fabricate behavioral arenas (Chowdhury et al., 2021; Gil-Martí et al., 2023). Setting up aggression assays (Dankert et al., 2009; Dierick, 2007) requires coating chambers with fluon to prevent flies from walking on the walls, which is tedious. Recording the videos require expensive machine vision cameras and machine vision backlights and technical expertise for data acquisition and processing. To overcome these challenges, we devised the DANCE hardware (Figure 6), which provides an inexpensive, easy, scalable, and robust alternative to existing methods of recording Drosophila aggression and courtship behaviors.

DANCE hardware and recordings setup.

(A) DANCE aggression set up (B) 3D- rendered components of the aggression set up (C) DANCE courtship set up (D) 3D- rendered components of the courtship set up, male and female are separated on either side using X-ray film separator or ‘divider comb’. (E-G) Top and side views of the DANCE setup with a smartphone camera for recording and electronic tablet being used as a backlight.

The DANCE hardware consists of easy to find off-the-shelf components including transparent medicine tablet foils/blister packs of different diameters, which are repurposed to serve as arenas for recording behaviors; acrylic sheets; and paper tape (Figure 6A-D, Supplementary video 1-2). To substitute expensive machine vision cameras, we used widely available Android smartphone cameras for recording and expensive machine vision backlights were replaced with electronic tablets or smartphones running white light application, to provide uniform backlight/illumination source (Figure 6E-G). Aggression and courtship behaviors were recorded at 30 fps with 1080 pixel resolution. In the DANCE aggression hardware (Figure 6A-B, E) medicinal foil is slid over a base plate containing an apple juice agar food layer over which flies fight (Supplementary Video 1-2). For DANCE courtship chambers (Figure 6C-D, F-G), the medicine tablet foil is slit from the middle using a razor, through which a thin ‘separator comb’ is inserted, which is made using x-ray film. This ‘separator comb’ keeps the males and females separate until the beginning of the recording, when it is manually removed (Supplementary video 3-4).

Since the heat generated from the tablet/smartphone screen during recording can affect the aggression and courtship behaviors, we placed a transparent acrylic sheet with an air- gap of 4 mm on top of the backlight screen to facilitate heat dissipation (Figure 6E-G; Supplementary Video 2, 4), which proved to be crucial to obtain robust data (Figure 7).

Benchmarking DANCE hardware and testing various neurogenetic tools.

(A- B) Courtship behaviors recorded using a pre-existing (circular) and DANCE set up, from group-housed (GH) and single-housed (SH) flies, for (C-D) wing extension, (C) ***p<0.0010, n=23; (D) **p<0.0013, GH, n=22 and SH, n=26. (E-F) Attempted copulation, (E) ***p<0.0002, n=23; (F) ns, p<0.1907, GH, n=18 and SH, n=22. (G-H) Following, (G) ns, p>0.0959, n=23; (H) ns, p<0.6589, GH, n=22 and SH, n=26. (I-J) Circling, (I) *p<0.012, n=23; (J) **p<0.0021, GH, n=19 and SH, n=22. (K-L) Aggressive lunges recorded using a pre-existing (circular) and DANCE set up. (M-N) Lunges in SH flies compared to GH flies reared on food with yeast granules, (M) **p<0.0138, n=36; (N) **p<0.0372, n=40. (O) Effect of yeast extract food on aggressive behavior; ****p<0.0001, n=38. (P-Q) Genetic knockdown of the neuropeptide Drosulfakinin (Dsk) in insulin-producing neurons using dilp2-GAL4. (P) ns, p<0.0502, ns, p>0.9999, ****p<0.0001, **p<0.0040, n=35. (Q) ****p<0.0001, ns, p>0.9999, ****p<0.0001, *p>0.0210, n=30. (R) Optogenetic silencing of dopaminergic neurons by UAS-GtACR1 driven by TH-GAL4 driver, ns, p<0.0986, ns, p>0.9999, ****p>0.0001, **p>0.0012, **p>0.0013, n=24. (C-J and M-O) Mann-Whitney U test; (P-R) Kruskal-Wallis test with Dunn’s multiple comparisons.

Benchmarking DANCE hardware

We used our DANCE classifiers to quantify aggression and courtship behaviors and found that wild type flies display these behaviors at similar levels in DANCE hardware as in the existing set ups (Figure7, Supplementary video 5). We investigated the effects of social isolation vs. enrichment on aggression and courtship behaviors, since single-housing (SH) when compared to group housing (GH), increases courtship attempts (Dankert et al., 2009; Kim and Ehrman, 1998; Pan and Baker, 2014), and promotes aggression (Agrawal et al., 2020; Wang et al., 2008; Yadav et al., 2024). We found comparable performance between pre-existing courtship setup (Koemans et al., 2017) (Figure 7A) and the DANCE hardware (Figure 7B), both of which captured significant differences in multiple courtship behaviors in SH vs. GH males (Figure 7C-J).

Comparison of pre-existing set up for aggression (Dankert et al., 2009) (Figure 7K) with the DANCE hardware (Figure 7L), showed that both the set ups faithfully captured even low levels of aggressive lunges in the wild type SH flies (Figure 7M-N). Availability of food affects Drosophila aggression (Lim et al., 2014) and yeast in fly food was shown to interact with gut microbiome to affect male aggression (Jia et al., 2021). We therefore investigated if the type of yeast used in food would affect aggression. Substituting yeast granules in fly food with yeast extract powder reduced the baseline aggression when flies were reared and differentially housed on food with yeast extract powder (Figure 7O). To investigate the effect of genetic knockdown using RNAi, which is a widely used tool in Drosophila genetics, we downregulated neuropeptide Drosulfakinin (Dsk) in insulin-producing neurons using dilp2- GAL4. Consistent with a previous finding (Agrawal et al., 2020), aggressive lunges were significantly increased in SH males (Figure 7P-Q). Behavioral screens often utilize optogenetics to identify neuronal circuits responsible for aggression (Hoopfer et al., 2015; Wohl et al., 2023; Yadav et al., 2024). It was earlier shown that while specific dopaminergic neurons may play a role in aggression, constitutive inactivation of a larger set of dopaminergic neurons using TH-GAL4 led to unhealthy flies with defects in locomotion due to which they did not fight (Alekseyenko et al., 2013). We reasoned that modern optogenetic tools may overcome the limitations of constitutive silencing and sought to investigate the role of dopaminergic neurons using our DANCE set up. We carried out optogenetic silencing during 20 minute interactions, by driving green light sensitive anion channelrhodopsin- GtACR1 (Govorunova et al., 2015; Mohammad et al., 2017) using TH- GAL4, and found significant increase in aggressive lunges in SH flies (Figure 7R). We also observed higher wing flicks as well as high intensity aggressive behaviors such as boxing and tussling (Supplementary Video 6). We did not find significant changes in locomotor activity between GH and SH flies despite 12 hours of continuous silencing (Figure 7— figure supplement 1). Together this suggests that the DANCE set up can be used with most of the common neurogenetic tools to screen for underlying molecular and neuronal circuitry governing aggression and courtship behaviors.

Discussion

Here, we present the DANCE assay, an easy to use and robust analysis pipeline and inexpensive hardware to record and quantify aggression and courtship behaviors. We developed six novel behavioral classifiers using supervised learning to accurately quantify aggression and courtship behaviors of Drosophila males. We developed a versatile, easy to fabricate and inexpensive hardware setup costing <$0.30, about 10,000 fold improvement over existing set ups, using easily available, off-the-shelf components. The performance of DANCE set up is comparable with more specialized and expensive existing setups. Therefore, DANCE enables rapid behavioral screening and wider adoption by the neuroscience community including in resource limited settings such as undergraduate education.

Various components of the DANCE assay including the behavioral classifiers, hardware design and analysis codes are publicly available and can be used independently as well. This offers flexibility to the neuroscience community to further customize the DANCE classifiers for specific needs or incorporate additional data without having to develop a classifier from scratch. Our study can also serve as a template for development of future behavioral classifiers for Drosophila aggression such as fencing, wing flick, box, tussle, chase, female headbutt etc. (Chen et al., 2002; Nilsen et al., 2004) and courtship behaviors, such as- tapping by male, licking the posterior abdomen of the female, female rejection etc. (Sokolowski, 2001).

Such high resolution analysis of complex social interactions could give us insights into mating dynamics, sexual selection and the influence of genetics and evolution. Further, the behavioral dynamics can help uncover sequencing of dynamic components of behaviors (Nilsen et al., 2004; Seeds et al., 2014; Simon and Heberlein, 2020; Zhang et al., 2020). It can also serve as a template for studying complex behaviors of various species of Drosophila and other insects. The ease of adaptability and portability of the DANCE assay can also be useful for ethologists studying insect behaviors closer to their natural habitat.

Materials and Methods

The details of all the custom codes, analysis pipeline, sample files to run the analysis and DANCE classifiers are available in our GitHub repository at- https://github.com/agrawallab/DANCE.

Fly Husbandry

Flies were reared on standard food at 25°C and 65% relative humidity with a 12hr:12hr light-dark cycle. All assays were performed at 25°C with 65% relative humidity, unless mentioned otherwise. For aggression and courtship experiments, Canton-S male flies were collected within 24 hours of eclosion and housed in a group (20 male flies per vial, 90mm length and 25mm diameter) or isolated (1 male fly per vial, 70mm length and 10mm diameter) for 6 days.

The following fly lines were acquired from the Bloomington Drosophila Stock Center (BDSC), USA: TH-GAL4, Dilp2-GAL4 (RRID:BDSC_37516), Dsk-RNAi (RRID:BDSC_25869) and attP2 empty vector control (RRID:BDSC_36303). The UAS- GtACR1 flies were a gift from Gaurav Das, NCCS, Pune, India and Canton-S (CS) flies were obtained from Ulrike Heberlein, HHMI, Janelia Research Campus, USA.

Aggression assay

The traditional aggression assay was performed as described previously (Dankert et al., 2009; Dierick, 2007). In brief, the behavioral chamber is made up of 12 well-aggressive arenas (10 mm in height and 16 mm in diameter/arena). These arenas were covered by a sliding lid with 2mm loading holes to facilitate the introduction of flies. A pair of male flies that were housed either as a group (GH) or single fly (SH) were introduced into the arena wells by gentle mouth aspiration through the loading holes. After the flies were loaded, the sliding lid was tightened using screws. Fluon (Insect-a-slip, Bioquip: cat #2871B) was coated on the arena walls and was allowed to dry overnight to create a slippery surface, thereby preventing the flies from climbing the walls. Sigmacote (Sigma-Aldrich: SL2) was used to coat the sliding lid of the arena to prevent flies from walking on the arena ceiling. The chamber was placed on the food plate containing commercial apple juice (without added sugars), 2.5% w/v of sucrose, and 2.25% w/v of agarose. For experiments to test effects of fly food nutrients, either 2.4% yeast extract powder (HiMedia: RM0271) or 2.4% Yeast granules (Prime Instant Dry yeast, AB Maury, India) were mixed in fly food. For optogenetic silencing experiments, 520-540 nm wavelength of LEDS (Lumileds:2835) were used at 0.0004 µW intensity, measured using a Power meter (Newport:843-R). For all aggression assays, flies were allowed to acclimatize in the arena for 5 minutes, and then the activity was filmed for 20 minutes. The assays were performed during ZT0 - ZT2.5 i.e. first 2.5 hours of the morning activity peak.

Courtship assay

Traditional courtship assays were performed as described by (Koemans et al., 2017; Ribeiro et al., 2018). We introduced single pairs of male and female flies using an aspirator in an 18-well courtship chamber, consisting of individual wells/arenas with a diameter of 10 mm. Male and female flies were introduced into one half of the chamber using the sliding entry holes with a removable separator that divided the chamber into two halves. The flies were allowed to acclimatize to the arena for 5 minutes, after which the separator was removed, and the courtship behavior was recorded for 15 minutes at 30 fps using a white backlight. The assay was performed from ZT0 to ZT3 or ZT9 to ZT12 (during peak activity windows). For generating mated females, we housed 20 females with 10 males together in a single vial for 4-6 days. For decapitated virgins, 2-4 days old virgin females were collected in a vial, and the decapitation was performed after anesthetizing on a CO2 fly-pad, just before the assay. The assays were performed during ZT0-ZT2.5 i.e., first 2.5 hours/morning activity peak or ZT9-ZT11.5 i.e. last 2.5 hours/evening activity peak.

MateBook

To quantify courtship data, an automated software called MateBook was used (https://github.com/Dicksonlab/MateBook) (Ribeiro et al., 2018). MateBook tracks the fly’s location and position for each video frame using the machine-vision approach. The software has built-in classifiers for each of the behaviors of a male fly during courtship, such as following, wing extension, orientation, copulation, and circling. The quantified data is generated in the form of a .tsv file, which contains bouts and other parameters related to courtship behavior. The software also provides an ethogram, which is a color-coded timeline of all the behaviors.

DANCE hardware

Circular, transparent medicinal tablet foils/blister packs were used to serve as aggression or courtship arenas, and acrylic sheets were used as base plates, and the DANCE hardware setup is held together using paper tape. The DANCE aggression arena was 13 mm in diameter and 4 mm in height. A thin acrylic base plate (2 mm thickness) with a loading hole (2 mm diameter) and a base plate with food and a side spacer were held together using paper tape (Supplementary Video 1). The tip of the food plate was also covered using paper tape for easy sliding of the medicinal tablet foil without damaging the food layer .

Sigmacote (Sigma-Aldrich: SL2) was used to coat the walls and roof of the medicine blister foil using cotton swab (Solimo, Amazon India), to prevent flies from climbing the walls of the arena. A pair of male files were introduced into the medicinal foil wells by gentle aspiration through the loading holes. After loading flies into the medicinal foil wells, the foil was slid onto the food bed with the help of side spacers and tape on the tip of the food bed. After sliding it, the assembly was taped from ends to avoid gaps and prevent flies from escaping the arenas (Supplementary Video 2).

The DANCE courtship arena used transparent medicine blister packs with a diameter of 11mm and a height of 4.5 mm. The top surface across all arenas (5 in a row) had thin slits cut using a sharp razor to fit in the separator comb which was built by cutting used x-ray film. The foil and base plate (2 mm thickness) with loading hole (2 mm diameter) was held together by a paper tape. The base plate for courtship had two holes spaced apart in such a way that, male and female flies can be loaded on either side of the separator comb. The assembly was taped from ends to avoid gaps and prevent flies from escaping the arenas.

Once the setup was ready to record, the separator comb was gently lifted up, while ensuring the slit edges stay intact to avoid distortions during recordings (Supplementary videos 3-4).

Video acquisition

For the traditional set ups, the interaction of the flies was recorded using machine-vision cameras (DMK 33UX252 USB 3.0 monochrome camera). A white backlight (TMS, BHS4- 00-100-X-W-24V) provided the light source for both the courtship and aggression experiments. Videos were recorded at 30 frames per second (fps) for 15 minutes for courtship or 20 minutes for aggression in H.264 (.mp4) format with 1440x1080 resolution. These videos were used for training, testing, and validating the DANCE classifiers.

Alternatively, for DANCE hardware testing, we used various Android smartphone cameras (Huawei Y9 2019; OnePlus Nord CE 2 Lite 5G, model: CPH2381 or Redmi Note 11 Pro+ 5G, model: 221116SI) at 30 fps with 1080p resolution in H.264 (.mp4) format for 15 minutes (courtship) or 20 minutes (aggression) with electronic tablet (iPad Air, 5th Generation) or smartphone (iPhone 13) running a ‘white screen light app’ as the background illumination source. A transparent acrylic sheet with 4 mm spacers was kept on top of this ‘backlight’ to create an air gap to ensure heat exchange and to prevent the DANCE arenas from getting hot.

Tracking flies using FlyTracker

To successfully classify the behavior of the flies, we tracked their location and body parts using Caltech FlyTracker (Eyjolfsdottir et al., 2014). It allows for detection and tracking of flies based on their location, body position, orientation, and interactions across the frames. Occurrence of identity switches between male and female flies was corrected using the ‘identity correction’ tool available in the “visualizer” program of the FlyTracker package. This data was then pushed to the JAABA pipeline to develop DANCE classifiers.

Developing the DANCE behavioral classifiers

The Janelia Automatic Animal Behavior Annotator (JAABA) was used to develop DANCE behavioral classifiers. The algorithm allows experienced users to ‘encode their intuition’ to develop ‘classifiers’ that help to annotate various animal behaviors using tracking data from the behavioral recordings (Kabra et al., 2013).

Tracking data generated using FlyTracker from training videos were processed using JAABA. The stereotypic behavior bouts were labeled as ‘true behavior’, e.g., ‘Lunge’, ‘Wing extension’ etc., and an approximately equal number of obvious non-behavior bouts (‘None’), were labeled. After each round of training, the false positives predicted by the classifier would be specifically labeled as ‘None.’ This cycle of training and correcting the predictions was done until there was no improvement in the classifier’s performance. We included courtship videos of males with decapitated virgins in the training set for the attempted copulation and circling classifiers to increase the number of labeled bouts of True behavior, as these behaviors occur more frequently with decapitated virgins. Additionally, we included videos with decapitated virgins in the training sets for the following and copulation classifiers to provide more examples of the ’None’ behavior. For the wing extension classifier, we used only mated females, as there were an adequate number of bouts representing both the true behavior and the ’None’ behavior.

Various DANCE classifiers were developed using following frames: lunge: 2730 frames (91 seconds in total); wing extension: 99,947 frames (3,331.57 seconds in total); attempted- copulation: 39,513 frames (1,317.1 seconds in total); copulation: 56,979 frames (1899.3 seconds in total); circling: 14,396 frames (479.87 seconds in total); following: 25,787 frames (859.57 seconds in total).

DANCE classifiers, custom codes and other analysis details are available in our GitHub repository at- https://github.com/agrawallab/DANCE.

Characterization of male aggression and courtship behaviors

Lunge: It is defined as ‘the attacking fly rises on his hind legs, lifting his long body axis by 45° then snaps down on his opponent with his head reaching a velocity of 200 mm/s’ (Dankert et al., 2009).

Courtship classifiers were developed for five male courtship behaviors based on existing definitions (Dankert et al., 2009; Ribeiro et al., 2018). For the duration-based classifiers, a filter of minimum bout length was added in the post-processing of each classifier as a measure to eliminate the bouts shorter than 98% of manual bouts. Details of individual courtship classifiers are mentioned below.

Wing Extension: Dankert et.al 2009, Ribeiro et al., 2018 define a wing extension, when the angle between the body’s main axis and the line from the body’s center to the tip of the wing exceeds 30°, with the behavior persisting for approximately 13 frames i.e. 0.5s, at 25 fps. We processed our data using MateBook as described (Ribeiro et al., 2018); but with a reduced persistence filter for wing extension to 0.33 seconds from the default 0.50 seconds to match our classifier’s threshold of minimum 10 frames, since in all the courtship videos that we analyzed, >98% of the bouts followed this threshold.

Attempted copulation and copulation: Existing definitions, such as (Ribeiro et al., 2018) define copulation as ‘occlusions’ persisting for more than 45 seconds and does not quantify ‘attempted-copulation’. We used a similar threshold while training our classifiers but ensured that it is not merely an ‘occlusion’— if the male curls it’s abdomen without mounting or mounts the female for a minimum of 0.33 seconds (10 frames at 30 fps) to a maximum of <45 seconds, it was considered as ‘attempted-copulation’ whereas mounting for >45 seconds (>1350 frames at 30 fps) was considered as ‘copulation’. The results were compared with MateBook by matching its ‘event settings’.

Circling: (Dankert et al., 2009) define circling as the fly drifting sideways in a circle with approximately constant velocity. Additionally, (Ribeiro et al., 2018) considered this behavior with a persistence of 13 frames (0.5s at 25 fps). We developed our ‘DANCE circling classifier’ based on these existing definitions, but with a reduced persistence of 10 frames (0.33s at 30 fps) and compared the results with MateBook by matching its ‘event settings’.

Following: Ribeiro et al., 2018 define the event following as when the ‘the male keeps distance to the female between 2 and 5mm, while directly behind the female, while both flies are walking with a minimum speed of 2mm/s’ with the behavior persisting for 25 frames i.e. 1.0s, at 25 fps. We developed our ‘DANCE following extension classifier’ based on these existing definitions, but with a reduced persistence of 10 frames (0.33s, at 30 fps) and compared the results with MateBook by matching its ‘event settings’.

Manual behavioral annotations

To quantitatively measure the classifier’s performance, manual behavioral annotations, or ‘ground-truthing’ were performed using the JAABA’s ground-truthing mode (Kabra et al., 2013). The classifier’s performance is measured by its robustness in predicting the behavioral bouts for the frames it was not trained on, which comprises the ‘testing set.’ In the testing set, we first annotate the behavioral bouts as ‘true behavior’ without looking at the predictions of the output. Then these videos are processed with the same classifier using JAABAPlot and compared as described subsequently.

Comparison of manual and DANCE annotations

The classifier’s output was compared with that of the manual ground-truth, based on both the number of bouts for the one-frame lunge classifier and the bout duration for all duration-based courtship classifiers (wing extension, following, circling, attempted-copulation, and copulation). The duration of the bouts was used to derive the respective behavioral index. A regression analysis and classification metrics were calculated between the bout numbers or the behavior index based on the type of classifier for each of the ground-truth, DANCE, and the existing software’s scores using Prism 8 (GraphPad Software).

The manual scores and JAABA output were compared, bout by bout, for the total number of true positives, false positives, and false negatives. Human annotated bouts (ground- truth) were used as a reference to the number of true positives. A ‘true positive’ is called when a JAABA bout overlaps with ground-truth bout for one or more frames (over 0.03 ms). If one ground-truth bout overlapped with two or more JAABA bouts, then all the JAABA bouts would be collectively counted as one bout, i.e., one true positive bout. Similarly, if two or more ground-truth bouts overlapped with a single JAABA bout, then the number of ground-truth bouts would be the count for the JAABA bout; i.e., two or more number of true positives.

When a JAABA bout does not overlap with any ground-truth (human annotation), then it would be considered a false positive. When a ground-truth bout does not overlap with any JAABA bout, it is considered a false negative (Leng et al., 2020).

We used a custom Python code, to calculate the number of total overlaps between the ground-truth, the DANCE classifier’s, and the existing algorithm’s output. Once we had the total number of true positives, false positives, and false negatives, we derived the precision, recall and F1 scores to gain a deeper understanding of the performance of DANCE classifiers. Precision rate represents the ratio of correctly predicted positive observations to the total predicted positives; the recall rate represents the ratio of correctly predicted positive observations to all actual positives; and the F1 score is a single metric which is a harmonic mean of both precision and recall. We used the following formulas to calculate precision, recall, and F1 score:

Statistical analysis

For non-normally distributed data, Mann-Whitney U test or ANOVA with appropriate posthoc correction was used. Statistical analysis of behavioral data was performed using Prism 8 (GraphPad Software), custom scripts or Microsoft Excel.

eLife Assessment

This study presents a valuable open-source and cost-effective method for automating the quantification of male aggression and courtship in Drosophila melanogaster. The work as presented provides solid evidence that the use of the behavioral setup that the authors designed - using readily available laboratory equipment and standardised high-performing classifiers they developed using existing software packages - accurately and reliably characterises social behavior in Drosophila. The work will be of interest to Drosophila neurobiologists and particularly to those working on male social behaviors.

Reviewer #1 (Public review):

The study introduces an open-source, cost-effective method for automating the quantification of male social behaviors in Drosophila melanogaster. It combines machine-learning-based behavioral classifiers developed using JAABA (Janelia Automatic Animal Behavior Annotator) with inexpensive hardware constructed from off-the-shelf components. This approach addresses the limitations of existing methods, which often require expensive hardware and specialized setups. The authors demonstrate that their new "DANCE" classifiers accurately identify aggression (lunges) and courtship behaviors (wing extension, following, circling, attempted copulation, and copulation), closely matching manually annotated ground-truth data. Furthermore, DANCE classifiers outperform existing rule-based methods in accuracy. Finally, the study shows that DANCE classifiers perform as well when used with low-cost experimental hardware as with standard experimental setups across multiple paradigms, including RNAi knockdown of the neuropeptide Dsk and optogenetic silencing of dopaminergic neurons.

The authors make creative use of existing resources and technology to develop an inexpensive, flexible, and robust experimental tool for the quantitative analysis of Drosophila behavior. A key strength of this work is the thorough benchmarking of both the behavioral classifiers and the experimental hardware against existing methods. In particular, the direct comparison of their low-cost experimental system with established systems across different experimental paradigms is compelling. While JAABA-based classifiers have been previously used to analyze aggression and courtship (Tao et al., J. Neurosci., 2024; Sten et al., Cell, 2023; Chiu et al., Cell, 2021; Isshi et al., eLife, 2020; Duistermars et al., Neuron, 2018), the demonstration that they work as well without expensive experimental hardware opens the door to more low-cost systems for quantitative behavior analysis.

Although the study provides a detailed evaluation of DANCE classifier performance, its conclusions would be strengthened by a more comprehensive analysis. The authors assess classifier accuracy using a bout-level comparison rather than a frame-level analysis, as employed in previous studies (Kabra et al., Nat Methods, 2013). They define a true positive as any instance where a DANCE-detected bout overlaps with a manually annotated ground-truth bout by at least one frame. This criterion may inflate true positive rates and underestimate false positives, particularly for longer-duration courtship behaviors. For example, a 15-second DANCE-classified wing extension bout that overlaps with ground truth for only one frame would still be considered a true positive. A frame-level analysis performance would help address this possibility.

In summary, this work provides a practical and accessible approach to quantifying Drosophila behavior, reducing the economic barriers to the study of the neural and molecular mechanisms underlying social behavior.

Reviewer #2 (Public review):

Summary:

This manuscript addresses the development of a low-cost behavioural setup and standardised open-source high-performing classifiers for aggression and courtship behaviour. It does so by using readily available laboratory equipment and previously developed software packages. By comparing the performance of the setup and the classifiers to previously developed ones, this study shows the classifier's overperformance and the reliability of the low-cost setup in recapitulating previously described effects of different manipulations on aggression and courtship.

Strengths:

The newly developed classifiers for lunges, wing extension, attempted copulation, copulation, following, and circling, perform better than available previously developed ones. The behavioural setup developed is low cost and reliably allows analysis of both aggression and courtship behaviour, validated through social experience manipulation (social isolation), gene knock (Dsk in Dilp2 neurons) and neuronal inactivation (dopaminergic neurons) known to affect courtship and aggression.

Weaknesses:

Aggression encompasses multiple defined behaviours, yet only lunges were analysed. Moreover, the CADABRA software to which DANCE was compared analyses further aggression behaviours, making their comparisons incomplete. In addition, though DANCE performs better than CADABRA and Divider in classifying lunges in the behavioural setup tested, it did not yield very high recall and F1 scores.

DANCE is of limited use for neuronal circuit-level enquiries, since mechanisms for intensity and temporally controlled optogenetic manipulations, which are nowadays possible with open-source software and low-cost hardware, were not embedded in its development.

Reviewer #3 (Public review):

The preprint by Yadav et al. describes a new setup to quantify a number of aggression and mating behaviors in Drosophila melanogaster. The investigation of these behaviors requires the analysis of a large number of videos to identify each kind of behavior displayed by a fly. Several approaches to automatize this process have been published before, but each of them has its limitations. The authors set out to develop a new setup that includes very low-cost, easy-to-acquire hardware and open-source machine-learning classifiers to identify and quantify the behavior.

Strengths:

(1) The study demonstrates that their cheap, simple, and easy-to-obtain hardware works just as well as custom-made, specialized hardware for analyzing aggression and mating behavior. This enables the setup to be used in a wide range of settings, from research with limited resources to classroom teaching.

(2) The authors used previously published software to train new classifiers for detecting a range of behaviors related to aggression and mating and to make them freely available. The classifiers are very positively benchmarked against a manually acquired ground truth as well as existing algorithms.

(3) The study demonstrates the applicability of the setup (hardware and classifiers) to common methods in the field by confirming a number of expected phenotypes with their setup.

Weaknesses:

(1) When measuring the performance of the duration-based classifiers, the authors count any bout of behavior as true positive if it overlaps with a ground-truth positive for only 1 frame - despite the minimal duration of a bout is 10 frames, and most bouts are much longer. That way, true positives could contain cases that are almost totally wrong as long there was an overlap of a single frame. For the mating behaviors that are classified in ongoing bouts, I think performance should be evaluated based on the % of correctly classified frames, not bouts.

(2) In the methods part, only one of the pre-existing algorithms (MateBook), is described. Given that the comparison with those algorithms is a so central part of the manuscript, each of them should be briefly explained and the settings used in this study should be described.

Taken together, this work can greatly facilitate research on aggression and mating in Drosophila. The combination of low-cost, off-the-shelf hardware and open-source, robust software enables researchers with very little funding or technical expertise to contribute to the scientific process and also allows large-scale experiments, for example in classroom teaching with many students, or for systematic screenings.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public review):

The study introduces an open-source, cost-effective method for automating the quantification of male social behaviors in Drosophila melanogaster. It combines machine-learning-based behavioral classifiers developed using JAABA (Janelia Automatic Animal Behavior Annotator) with inexpensive hardware constructed from off-the-shelf components. This approach addresses the limitations of existing methods, which often require expensive hardware and specialized setups. The authors demonstrate that their new "DANCE" classifiers accurately identify aggression (lunges) and courtship behaviors (wing extension, following, circling, attempted copulation, and copulation), closely matching manually annotated groundtruth data. Furthermore, DANCE classifiers outperform existing rule-based methods in accuracy. Finally, the study shows that DANCE classifiers perform as well when used with low-cost experimental hardware as with standard experimental setups across multiple paradigms, including RNAi knockdown of the neuropeptide Dsk and optogenetic silencing of dopaminergic neurons.

The authors make creative use of existing resources and technology to develop an inexpensive, flexible, and robust experimental tool for the quantitative analysis of Drosophila behavior. A key strength of this work is the thorough benchmarking of both the behavioral classifiers and the experimental hardware against existing methods. In particular, the direct comparison of their low-cost experimental system with established systems across different experimental paradigms is compelling.

While JAABA-based classifiers have been previously used to analyze aggression and courtship (Tao et al., J. Neurosci., 2024; Sten et al., Cell, 2023; Chiu et al., Cell, 2021; Isshi et al., eLife, 2020; Duistermars et al., Neuron, 2018), the demonstration that they work as well without expensive experimental hardware opens the door to more low-cost systems for quantitative behavior analysis.

We thank the reviewer for their positive assessment and constructive suggestions. We have cited these additional JAABA studies in the Introduction. We clarified that several prior JAABA-based classifiers were developed using specialized machinevision cameras or custom setups, and that in some cases the original code and classifiers were not made publicly available, which limits reproducibility and wider adoption. To address this, we explicitly note in the revised manuscript that DANCE was developed with accessibility in mind.

Although the study provides a detailed evaluation of DANCE classifier performance, its conclusions would be strengthened by a more comprehensive analysis. The authors assess classifier accuracy using a bout-level comparison rather than a frame-level analysis, as employed in previous studies (Kabra et al., Nat Methods, 2013). They define a true positive as any instance where a DANCE-detected bout overlaps with a manually annotated ground-truth bout by at least one frame. This criterion may inflate true positive rates and underestimate false positives, particularly for longer-duration courtship behaviors. For example, a 15-second DANCE-classified wing extension bout that overlaps with ground truth for only one frame would still be considered a true positive. A frame-level analysis performance would help address this possibility.

We thank the reviewer for raising this important point. Our original use of bout-level analysis followed existing literature (Duistermars et al., 2018; Ishii et al., 2020; Chiu et al., 2021; Tao et al., 2024; Hindmarsh Sten et al., 2025). While our lunge classifier already operates at the frame level, we have now performed additional frame-level evaluations for the duration based courtship classifiers. These analyses revealed only minor differences in precision, recall, and F1 scores compared with the original bout-level approach (see new Figure 5—Figure Supplement 3). Details of this analysis are now included in the Materials and Methods.

In summary, this work provides a practical and accessible approach to quantifying Drosophila behavior, reducing the economic barriers to the study of the neural and molecular mechanisms underlying social behavior.

We thank the reviewer for their encouraging comments and for recognizing the accessibility and practical value of our approach. We appreciate the constructive suggestions, which have helped strengthen the manuscript.

Reviewer #2 (Public review):

Summary:

This manuscript addresses the development of a low-cost behavioural setup and standardised open-source high-performing classifiers for aggression and courtship behaviour. It does so by using readily available laboratory equipment and previously developed software packages. By comparing the performance of the setup and the classifiers to previously developed ones, this study shows the classifier's overperformance and the reliability of the low-cost setup in recapitulating previously described effects of different manipulations on aggression and courtship.

Strengths:

The newly developed classifiers for lunges, wing extension, attempted copulation, copulation, following, and circling, perform better than available previously developed ones. The behavioural setup developed is low cost and reliably allows analysis of both aggression and courtship behaviour, validated through social experience manipulation (social isolation), gene knock (Dsk in Dilp2 neurons) and neuronal inactivation (dopaminergic neurons) known to affect courtship and aggression.

We thank the reviewer for the clear summary of our work and for highlighting its strengths. We appreciate these positive comments and suggestions, which have helped improve the clarity of the manuscript.

Weaknesses:

Aggression encompasses multiple defined behaviours, yet only lunges were analysed. Moreover, the CADABRA software to which DANCE was compared analyses further aggression behaviours, making their comparisons incomplete. In addition, though DANCE performs better than CADABRA and Divider in classifying lunges in the behavioural setup tested, it did not yield very high recall and F1 scores.

We thank the reviewer for raising this important point. We focused on lunges because they are widely used as a standard proxy for male aggression across multiple laboratories (Agrawal et al., 2020; Asahina et al., 2014; Chiu et al., 2021; Chowdhury et al., 2021; Dierick et al., 2007; Hoyer et al., 2008; Jung et al., 2020; Nilsen et al., 2004; Watanabe et al., 2017). As noted in the Discussion, our study also provides a template for the future development of additional aggression classifiers (fencing, wing flick, tussle, chase, female headbutt) and courtship classifiers (tapping, licking, rejection), which can be trained and shared through the same DANCE framework. Developing and validating these was beyond the scope of the present work.

To address the concern regarding precision, recall, and F1 scores, we performed additional analyses across all training videos and compiled these results in the new Figure 2—Figure Supplement 2. Our earlier lunge classifier had performance metrics obtained after training on a total of 11 videos. Our analysis shows performance metrics for classifiers trained on four independent datasets (Videos 8– 11). We found that the classifier trained on nine videos provided the best balance of precision, recall, and F1 (78.73%, 73.07%, and 75.79%, respectively), which was slightly better than the earlier classifier. We therefore updated the main figure, text, and Materials and Methods to use this version and uploaded the corresponding classifier and training details to the GitHub repository.

DANCE is of limited use for neuronal circuit-level enquiries, since mechanisms for intensity and temporally controlled optogenetic manipulations, which are nowadays possible with open-source software and low-cost hardware, were not embedded in its development.

We thank the reviewer for this valuable point. The primary aim of DANCE is to provide an accessible, modular, and low-cost behavioural recording and analysis platform. It was designed so that users can readily integrate additional components such as optogenetic control when needed. As a proof of concept, we implemented optogenetic silencing of dopaminergic neurons using the DANCE hardware and confirmed that this manipulation increased aggression (Figure 7R).

To facilitate adoption, we now provide schematic diagrams, LED control code, and instructions on our GitHub page and setup photographs in the manuscript (see new Figure 7—Figure Supplement 1). The released code allows programmable timing and intensity control, enabling users to reproduce temporally precise optogenetic protocols or extend the system for other stimulation paradigms.

Reviewer #3 (Public review):

The preprint by Yadav et al. describes a new setup to quantify a number of aggression and mating behaviors in Drosophila melanogaster. The investigation of these behaviors requires the analysis of a large number of videos to identify each kind of behavior displayed by a fly. Several approaches to automatize this process have been published before, but each of them has its limitations. The authors set out to develop a new setup that includes very low-cost, easy-to-acquire hardware and open-source machine-learning classifiers to identify and quantify the behavior.

Strengths:

(1) The study demonstrates that their cheap, simple, and easy-to-obtain hardware works just as well as custom-made, specialized hardware for analyzing aggression and mating behavior. This enables the setup to be used in a wide range of settings, from research with limited resources to classroom teaching.

(2) The authors used previously published software to train new classifiers for detecting a range of behaviors related to aggression and mating and to make them freely available. The classifiers are very positively benchmarked against a manually acquired ground truth as well as existing algorithms.

(3) The study demonstrates the applicability of the setup (hardware and classifiers) to common methods in the field by confirming a number of expected phenotypes with their setup.

We thank the reviewer for the positive assessment of our work and for highlighting its strengths. We appreciate these encouraging comments and suggestions, which have helped improve the clarity and presentation of the manuscript.

Weaknesses:

(1) When measuring the performance of the duration-based classifiers, the authors count any bout of behavior as true positive if it overlaps with a ground-truth positive for only 1 frame - despite the minimal duration of a bout is 10 frames, and most bouts are much longer. That way, true positives could contain cases that are almost totally wrong as long there was an overlap of a single frame. For the mating behaviors that are classified in ongoing bouts, I think performance should be evaluated based on the % of correctly classified frames, not bouts.

We thank the reviewer for raising this concern. In response to this point, and to Reviewer #1’s similar comment, we performed a frame-level evaluation of all duration-based courtship classifiers. The analysis revealed only minor differences compared with the original bout-level metrics (see new Figure 5—Figure Supplement 3), confirming the robustness of our classifiers. We have also added a description of this analysis in the Materials and Methods section.

(2) In the methods part, only one of the pre-existing algorithms (MateBook), is described. Given that the comparison with those algorithms is a so central part of the manuscript, each of them should be briefly explained and the settings used in this study should be described.

We thank the reviewer for this helpful suggestion. In the revised manuscript, we expanded the Materials and Methods to include concise descriptions and parameter settings for all pre-existing algorithms used for comparison. This includes dedicated subsections for CADABRA and the Divider assay, with explicit reference to their rulebased or geometric features. For MateBook, we specified the persistence filters used and the adjustments made for fair benchmarking. These changes ensure transparency and reproducibility.

Taken together, this work can greatly facilitate research on aggression and mating in Drosophila. The combination of low-cost, off-the-shelf hardware and open-source, robust software enables researchers with very little funding or technical expertise to contribute to the scientific process and also allows large-scale experiments, for example in classroom teaching with many students, or for systematic screenings.

We thank the reviewer for the encouraging comments and for recognizing the accessibility and broad applicability of DANCE. We believe these revisions have further strengthened the manuscript.

Reviewer #1 (Recommendations for the authors):

The following comments highlight areas where additional context, clarification, or further analysis could strengthen the manuscript. I hope these suggestions will be useful in refining your work.

(1) Lines 71-73: The authors state that Ctrax "leads to frequent identity switches among tracked flies, which is not the case while using FlyTracker." However, Ctrax was specifically designed to minimize identity errors, and Kabra et al. (2013) reported a low frequency of such errors-approximately one per five fly-hours in 10-fly videos. In contrast, Caltech FlyTracker does not correct identity errors automatically, requiring manual corrections, as noted in the Methods section of this study. If this is not an oversight, please provide further context to clarify this distinction.

We thank the reviewer for raising this clarification. As reported by Bentzur et al. (2021), when groups of flies were tracked simultaneously, Ctrax often generated multiple identities for the same individual, sometimes producing more trajectories than the actual number of flies. To prevent ambiguity, we revised the text to read: “While both Ctrax and FlyTracker (Eyjolfsdottir et al., 2014) may produce identity switches, when groups of flies were tracked simultaneously, Ctrax led to inaccuracies that required manual correction using specialized algorithms such as FixTrax (Bentzur et al., 2021).” We also quantified FlyTracker identity-switch rates in our datasets and report them in new Supplementary File 5, confirming that such events were rare (< 2% of tracked intervals). We believe, this updated version provides the necessary context and ensures accuracy in describing each tracker’s limitations.

(2) Line 85: Providing additional context on how this study builds on previous work using JAABA-based classifiers for fly social behavior and comparing these classifiers to rule-based methods would more accurately situate it within the field. The authors state that "recently, a few JAABA-based classifiers have been developed for measuring aggression and courtship" and cite four related studies. However, this statement seems to underrepresent the use of JAABA-based classifiers for quantifying fly social behavior, which has become common in the field. Several additional studies (as noted in the public review) have developed JAABA-based classifiers for scoring aggression or courtship. Furthermore, other studies have compared the performance of JAABA-based classifiers with rule-based classifiers like CADABRA (e.g., Chowdhury et al., Comm Biology 2021; Leng et al., PlosOne 2020; Kabra et al., Nat Methods 2013). Mentioning the similar findings in those studies and your own helps strengthen the conclusion that machine-learning-based classifiers outperform rule-based classifiers in several experimental contexts.

We thank the reviewer for this helpful suggestion. We have revised the Introduction to include additional references to studies that applied JAABA-based classifiers for aggression and courtship and made textual edits to reflect this. We further noted that, unlike several previous studies, all DANCE classifiers and analysis code are publicly available.

Reviewer #2 (Recommendations for the authors):

(1) Suggestions for improved or additional experiments, data or analyses: As mentioned in the description of the effect of optogenetic inactivation of dopaminergic neurons, in the conclusion and also reported in the literature, there are other important identified aggression behaviours, such as fencing, wing flick, tussle, and chase. Similarly, for courtship, tapping and licking have also been defined. This study, as opposed to proposed future studies, would benefit from creating opensource classifiers for these established behaviours, which are important for the analysis of aggression and courtship.

We thank the reviewer for this valuable suggestion. As clarified in the Discussion, this manuscript intentionally focuses on six core, well-validated aggression and courtship behaviors to demonstrate DANCE’s modularity and reproducibility. Developing additional classifiers such as fencing, wing flick, tussle, chase, tapping, and licking would require extensive annotation and validation beyond the present scope. To address this point, we explicitly note in the revised text that the DANCE pipeline is readily extendable, allowing the community to build new classifiers within the same framework.

In terms of observer bias assessment for ground-truthing in courtship, this was only presented for circling and it would be beneficial to have encompassed all behaviours analysed.

We thank the reviewer for this suggestion. Observer-bias comparisons for all six classifiers are presented in Figure 2—Figure Supplement 1 (panels A–F). We clarified in the Results that annotations from two independent evaluators were compared for all classifiers, with no significant differences observed, confirming their robustness.

Finally, intensity and temporal optogenetic control are important for neuronal circuit analysis of underlying behaviour. The authors could embed this aspect in DANCE by integrating control of the green light LED strip used in this study using, for example, the open-source visual reactive programming software Bonsai (Lopes et al., 2015) and open-source electronics platform Arduino. This is an important and valuable addition in line with maintaining low cost.

We thank the reviewer for this valuable suggestion. DANCE was designed to be modular, allowing integration of temporal optogenetic control. To support immediate adoption, we now provide Arduino LED control code, setup schematics, and photographs (new Figure 7—Figure Supplement 1) along with step-by-step instructions on our GitHub page. We also note that Bonsai and Arduino frameworks are compatible with DANCE, enabling future extensions for closed-loop or behaviortriggered stimulation.

(2) Minor corrections to the text and figures:

Figure Supplement 1 refers only to Figure 2, yet panels D-F refer to the behaviour circling in courtship and therefore should be assigned to the respective figure.

Thanks, we have corrected this.

In lines 315-316, the cumbersome task of fluon coating for aggression assays seems to be ubiquitous across assays which is not the case, and therefore the sentence should include the word 'some'.

Thanks, we have edited this.

The cost of the phone and/or tablet should be included in the DANCE setup costs, as presumably these devices will be dedicated to the behavioural studies, for consistency purposes.

We thank the reviewer for this comment. We intentionally did not include smartphones or tablets in the setup cost because, in our experiments, these devices were not dedicated exclusively to DANCE but were repurposed from routine personal use. Our aim was to leverage readily available consumer electronics so that their cost does not become a barrier to adoption. We confirmed that commonly available Android phones capable of 30 fps at 1080p in H.264 format, as well as tablets or phones running a simple white-screen light app, are sufficient for reliable behavior classification and illumination. Since these devices can be returned to regular use after recordings, including their cost in the setup would not accurately reflect the intended accessibility of DANCE. For consistency, we now clarify in the Materials and Methods that such devices should be placed in airplane mode during recordings.

Reviewer #3 (Recommendations for the authors):

(1) For my taste, the authors put too much emphasis on the point that their method outperforms existing methods. I understand the value in comparing to published methods and it is of course fully justified to state the advantages of the new method. But the whole preprint is set up as a competition with the old algorithms, and the conclusion that the new classifier is better is repeated in each figure caption and after each paragraph of the results. This competitive mindset also extends to the selection of which results are presented as main figures and which as supplements - all cases in which the previous methods actually perform well are only presented in the supplement. I think this is simply unnecessary as the authors' results speak for themselves, and do not need the continuous competitive comparison.

We thank the reviewer for this thoughtful suggestion. Our intention was to benchmark DANCE rigorously against existing methods, not to frame the study competitively. We agree that repeated emphasis on relative performance was unnecessary. In the revised version, we streamlined figure captions and text throughout the manuscript to balance comparisons and removed redundant phrasing. Instances where other methods performed well are now presented with equal clarity to maintain a neutral and informative tone.

(2) When describing the DANCE hardware, as a reader I would find it interesting to also read about potential issues that the authors encountered. For example, how difficult is it to handle the materials without breaking or deforming them, which could affect the behavioral assays? How critical is it to use specific blister packs - the availability of which will likely vary strongly between countries? Did the authors try different sizes, and products? Such information, even as a supplement, could be very helpful for the widespread use of the hardware.

We thank the reviewer for this important point. To address this, we conducted additional tests comparing DANCE arenas of different diameters (new Figure 7— Figure Supplement 3A–C and new Figure 7—Figure Supplement 4A–L). We also consulted colleagues in multiple countries and verified that the blister packs used in our assays are readily available. The Materials and Methods now include practical handling notes: blister foils can be reused ~30–40 times for aggression assays and ~10–15 times for courtship assays before deformation. We also describe how to prevent agar surface damage during assembly and how to wash and dry the arenas for optimal reusability.

(3) I find the arrows pointing to several videos in a number of figures rather distracting and redundant, and suggest omitting them.

Thanks, we have omitted these arrows from all relevant figures and clarified the figure legends to enhance readability.

(4) P8, line 169 ff: this is a very long sentence that should be separated into several sentences.

We have rewritten this as follows: “DANCE scores remained comparable to groundtruth scores across all categories, whereas CADABRA and Divider underestimated the lunge counts (Figure 2B–E). Correlation analysis revealed a strong relationship between DANCE and ground-truth scores (Figure 2F, Supplementary File 2). In comparison, CADABRA and the Divider assay classifier showed a weaker correlation (Figure 2G-H, Supplementary File 2).”

(5) P10, line 216: please explain, here and in the methods, how these behavioral indices are calculated. I did not find this information anywhere in the paper.

We thank the reviewer for pointing this out. We now define the behavioral index explicitly in Materials and Methods: “For each assay, a behavioral index was calculated as the proportion of frames in which the male engaged in the specified behavior. This was obtained by dividing the total number of frames annotated for that behavior by the total number of frames in the recording.”

(6) P11, line 253: I don't understand the modifications to MateBook regarding attempted copulations, neither in the results nor the methods section. I would ask the authors to explain more explicitly what was done.

We thank the reviewer for this helpful suggestion. We have re-written several parts of the Materials and methods to clarify these details and streamline the text. To train the attempted copulation classifier, we combined datasets from assays with mated and decapitated virgin females, using manual annotations as ground truth. We also adapted MateBook’s persistence filters (Ribeiro et al., 2018) and defined thresholds explicitly: mounting lasting >45 s (>1350 frames at 30 fps) was defined as copulation, whereas abdominal curling without mounting, or mounting lasting 0.33– 45 s, was defined as attempted copulation.

(7) Figure 7F: this is the only case with a significant difference between the two setups. What explanations do the authors have for the discrepancy?

We thank the reviewer for raising this point. After repeating the experiments, we no longer found a significant difference between the setups. Figure 7 and its legend have been updated to reflect these results.

(8) Figure 2 - Supplement 1: I do not understand why the boxes for Observer 1 have different colors in different figures. Does this have a meaning?

Thanks for pointing this out. The color differences had no intended meaning, and we have corrected the figure for consistency across panels.

(9) P22, line 517ff: It would be interesting to know how frequently identity switches occurred. For large-scale, automatic behavioral screenings that step could be a crucial bottleneck.

We thank the reviewer for this valuable suggestion. We analyzed identity switches using the FlyTracker “Visualizer” package, which flags frames with possible overlaps or jumps. Flagged intervals were manually verified, and we report these data in new Supplementary File 5. Identity switch rates were very low: 0.66% for high-resolution recordings and 1.9% for smartphone DANCE videos in the most challenging decapitated-virgin dataset. These findings demonstrate robust tracking performance under both setups.

Version 2

Abstract

Quantifying animal behavior is pivotal for identifying the underlying neuronal and genetic mechanisms involved. Computational approaches have enabled automated analysis of complex behaviors such as aggression and courtship in Drosophila. However, existing approaches rely on rigid, rule-based algorithms and expensive hardware, limiting sensitivity to behavioral variations and accessibility. Here, we describe the Drosophila Aggression and Courtship Evaluator (DANCE), a low-cost, open-source platform that combines machine learning-based classifiers and inexpensive hardware to quantify aggression and courtship. DANCE consists of six novel behavioral classifiers trained using a supervised machine learning algorithm. DANCE classifiers address key limitations of rule-based algorithms, capturing dynamic behavioral variations more effectively. DANCE hardware is constructed using repurposed medicine blister packs and acrylic sheets, with recordings performed using smartphones, making it affordable and accessible. Benchmarking demonstrated that DANCE hardware performs comparably to sophisticated, high-cost setups. We validated DANCE in diverse contexts, including social isolation versus enrichment, which modulates aggression and courtship; RNAi-mediated downregulation of the neuropeptide Dsk; and optogenetic silencing of dopaminergic neurons, which promotes aggression. DANCE provides a cost-effective and portable solution for studying Drosophila behaviors in resource-limited settings or near natural habitats. Its accessibility and robust performance democratize behavioral neuroscience, enabling rapid screening of genes and neuronal circuits underlying complex social behaviors.

Introduction

Detailed and accurate annotation and analysis of complex behaviors are necessary for understanding the underlying neural and molecular mechanisms. The fruit fly Drosophila melanogaster is one of the most accessible and well-studied model organisms for identifying the neuronal and molecular underpinnings of behavior. Multiple large-scale screens have been conducted in Drosophila to study complex social behaviors such as aggression and courtship (Asahina, 2017; Greenspan and Ferveur, 2000; Hall, 2002; Kravitz and Fernandez, 2015) to identify the underlying neural circuitry (Agrawal et al., 2020; Asahina et al., 2014; Davis et al., 2018; Hoopfer et al., 2015; Yadav et al., 2024) and genes involved (Agrawal et al., 2020; Benzer, 1967; Gill, 1963; Hall, 1978; Ishii et al., 2022; Wang et al., 2008). These behaviors exhibit distinct, stereotyped patterns. For example, aggression involves chasing, fencing (Jacobs, 1960), wing threats, boxing (Dow and Schilcher, 1975), lunging, and tussling (Hoffmann, 1987a, 1987b). Similarly, courtship consists of multiple stereotyped behaviors exhibited by the male fly, such as orienting, circling, and following the female (Cook and Cook, 1975; Markow, 1987; O’Dell, 2003). To stimulate the female to be more receptive, the male produces a species-specific song by vibrating and extending its wing (Bennet-Clark and Ewing, 1969; Swain and von Philipsborn, 2021). The male then attempts copulation by curling its abdomen and finally mounts the female for copulation (Bastock and Manning, 1955; Spieth, 1974).

Manual analysis by trained observers is considered the gold standard in behavioral analysis, but it is time-consuming and unsuitable for large-scale screens (Gomez-Marin et al., 2014; Robie et al., 2017a). ‘Computational ethology’ (Anderson and Perona, 2014; Datta et al., 2019) helps address this challenge by automating behavioral annotation by leveraging advances in computer vision and machine learning (Robie et al., 2017b). This enables high-throughput behavioral screening to identify responsible genes and circuits.

A typical computational ethology workflow involves recording animal behaviors and tracking their positions along with body movements. This is followed by the analysis and classification of the observed behaviors from hundreds to thousands of video frames capturing behavioral instances. Several software programs, such as Ctrax, Caltech FlyTracker, and Deep Lab Cut (Branson et al., 2009; Eyjolfsdottir et al., 2014; Mathis et al., 2018), are widely used for tracking behaviors in Drosophila. Each comes with strengths and weaknesses. Ctrax (Branson et al., 2009) can accurately track fly position and movement, but identity switches remain a challenge, especially when tracking groups of flies. While both Ctrax and FlyTracker (Eyjolfsdottir et al., 2014) may produce identity switches, when groups of flies were tracked simultaneously, Ctrax led to inaccuracies that required manual correction using specialized algorithms such as FixTrax (Bentzur et al., 2021).

The effectiveness of various machine learning pipelines is eventually measured by comparing their output to human annotation, called ‘ground-truthing’. Rule-based algorithm such as CADABRA (Dankert et al., 2009) is used to quantify aggression, but it can lead to mis-scoring and identity switches, as revealed by ground-truthing (Simon and Heberlein, 2020), which needs to be corrected in a semi-automated manner (Kim et al., 2018). MateBook (Ribeiro et al., 2018) is another rule-based algorithm used to quantify courtship; however, similar to CADABRA, it tends to miss true-positive events, leading to significant mis-scoring of behaviors under certain experimental conditions.

The Janelia Automatic Animal Behavior Annotator (JAABA) (Kabra et al., 2013) addresses the challenges of rigid rule-based approaches by employing a supervised learning approach. In the JAABA pipeline, user-labeled data are utilized for training to encompass the dynamic variations in behaviors, allowing it to predict behaviors on the basis of learning from input data.

Several studies have developed JAABA-based behavioral classifiers for measuring aggression (Chiu et al., 2021; Chowdhury et al., 2021; Duistermars et al., 2018; Leng et al., 2020; Tao et al., 2024) and courtship (Gil-Martí et al., 2023; Pantalia et al., 2023). However, many of these studies did not make these classifiers publicly available (Duistermars et al., 2018; Gil-Martí et al., 2023; Hindmarsh Sten et al., 2025; 2020; Pantalia et al., 2023). In other cases, the reported approaches relied on specialized hardware, such as custom 3D printed parts (Chowdhury et al., 2021; Gil-Martí et al., 2023), or high-end machine-vision cameras (Chiu et al., 2021; Chowdhury et al., 2021; Duistermars et al., 2018; Hindmarsh Sten et al., 2025; Leng et al., 2020; Tao et al., 2024), limiting their accessibility and wider adoption.

Here, we describe DANCE (Drosophila Aggression and Courtship Evaluator), an open-source, user-friendly analysis and hardware pipeline to simplify and automate the process of robustly quantifying aggression and courtship behaviors. DANCE has two components: 1. A set of robust, machine vision-based behavioral classifiers developed using JAABA to quantify aggression and courtship; and 2. An inexpensive hardware setup built from off-the-shelf materials and consumer smartphones for behavioral recording. Compared with previous methods (Dankert et al., 2009; Ribeiro et al., 2018), the DANCE classifiers improved accuracy and reliability, while its low-cost hardware eliminates the need for specialized arenas and cameras. All classifiers and analysis codes are publicly available, enabling broad adoption, especially in resource-limited settings. Together, DANCE provides a powerful, accessible platform for behavioral screening and the discovery of mechanisms underlying complex social behaviors and neurological disorders.

Results

DANCE assay analysis pipeline

To overcome the challenge of time-consuming manual behavioral annotation or resource-intensive, complex hardware, we developed an automated, high-throughput quantification pipeline—DANCE and trained new behavioral classifiers using an existing machine learning algorithm, JAABA (Kabra et al., 2013)—to robustly quantify aggression and courtship in Drosophila (Figure 1). We also designed a simple, low-cost recording setup constructed from repurposed transparent medicine blister packs, acrylic sheets, and paper tape, enabling easy behavioral recordings. To record these behaviors, we used Android smartphone cameras and an electronic tablet or smartphone serving as a backlight illumination source (Figure 1A, Materials and Methods).

The DANCE assay provides an accessible approach for quantifying aggression and courtship behaviors.

(A) Comparison of existing machine-vision camera hardware with the simplified, low-cost DANCE hardware for behavior acquisition. (B) Workflow for developing DANCE classifiers, including training, benchmarking against existing methods and manual ground-truthing to generate behavioral scores. (C) Behavioral classifiers developed to quantify male aggression (lunge) and courtship (wing extension, circling, following, attempted copulation, and copulation). (D) Representative raster plots comparing ground-truth, DANCE, CADABRA, and Divider assay performance for aggression. (E) Representative raster plots comparing ground-truth, DANCE, and MateBook performance for courtship. Created with BioRender.com/i1hyc7d.

We quantify these behaviors using our DANCE classifiers. Unlike existing set-ups that cost approximately USD 3500, DANCE hardware can be assembled from off-the-shelf components for less than USD 0.30 (Supplementary File 1). To benchmark the performance of the DANCE classifiers, we used pre-existing setups and rule-based methods for quantifying courtship and aggression and compared their performance with that of the DANCE classifiers (Figure 1B-C).

To train the DANCE classifiers using JAABA (Kabra et al., 2013), for aggressive lunges, we used an existing setup described in (Dankert et al., 2009) modified from (Dierick, 2007) (Figure 1A; Figure 1—Figure Supplement 1), and for courtship behaviors, we used a pre-existing setup described in (Koemans et al., 2017) (Figure 1A; Figure 1—Figure Supplement 2). We tracked the position, motion, and interactions of pairs of flies across video frames using the Caltech FlyTracker (Eyjolfsdottir et al., 2014). To avoid data leakage, we randomly divided the acquired videos into two categories, "training videos" and “test videos,” to train and evaluate the DANCE classifiers. These test videos were also manually ‘ground-truthed’ frame by frame, which is considered the gold standard for behavioral annotation (Figure 1B, D-E).

We benchmarked the performance of the DANCE classifiers against existing rule-based algorithms, CADABRA (Dankert et al., 2009) and MateBook (Ribeiro et al., 2018), and an existing JAABA aggression classifier (Chowdhury et al., 2021). Comparisons with manual ground-truth data revealed that the performance of the DANCE classifiers is comparable to that of human annotations and has higher sensitivity than rule-based algorithms (Figure 1D-E). The subsequent sections describe the quantitative analysis of individual DANCE classifiers and benchmarking of DANCE hardware.

DANCE lunge classifier to quantify aggressive behavior

Aggression is an innate, complex behavior, and Drosophila males exhibit several stereotyped behavioral patterns during aggressive encounters, with lunging used widely as a measure of overall aggression in males (Agrawal et al., 2020; Asahina et al., 2014; Chiu et al., 2021; Chowdhury et al., 2021; Davis et al., 2014; Dierick, 2007; Hoffmann, 1987a; Hoopfer et al., 2015; Hoyer et al., 2008; Jung et al., 2020; Nilsen et al., 2004; Watanabe et al., 2017; Yadav et al., 2024). A lunge is defined as a male fly raising its front legs and hitting down on the other fly.

We developed a new classifier using JAABA (Kabra et al., 2013) to robustly quantify aggressive lunges in Drosophila, hereafter referred to as the DANCE lunge classifier. We quantified lunges using our classifier from 20-minute long videos and compared the output with manual ground-truth and existing methods—CADABRA (Dankert et al., 2009) and the Divider assay classifier (Chowdhury et al., 2021). CADABRA tends to miss several true positive lunges, likely because of its rigid, rule-based framework, which cannot adapt to the dynamic variations in behavior. Figure 2A shows the lunge scores from 40 different videos using the ground-truth, the DANCE lunge classifier, CADABRA, and the Divider assay classifier. While the ground-truth and DANCE classifiers’ outputs are comparable, CADABRA and the Divider assay classifier underscore lunges across videos. We ground-truthed the DANCE lunge classifier against 40 ‘test videos’ (Figure 2A, Materials and Methods). Inter-observer validation confirmed that there were no significant differences between the two independent manual annotations (Figure 2—Figure Supplement 1A).

Validation of the DANCE lunge classifier for quantifying male aggression.

(A) Lunge scores from 20-minute-long videos obtained by the ground-truth (gray), the DANCE lunge classifier (orange), CADABRA (purple) and Divider assay classifier (green). (B-E) Comparison of lunge scores across different aggression levels, based on manual scoring and predictions from DANCE, CADABRA, and Divider: (B) 0–70 lunges ‘low aggressive’ (n=10; **P<0.0017, ****P<0.0001). (C) 71–160 lunges, ’moderately aggressive’, (n=11; **P<0.0102, ***P<0.0002, ***P<0.0001). (D) 161–300 lunges, ‘highly aggressive’, (n = 11; **P<0.0057, ***P<0.0002, ***P<0.0004). and (E) >300 lunges, ‘hyper-aggressive’, (n = 8; *P<0.0402, **P<0.0029, ***P<0.0006; Friedman’s ANOVA with Dunn’s test) (F) Regression analysis of the DANCE ‘lunge classifier’ vs. manual scores (R2=0.9760, n=40). (G) Regression of the CADABRA vs. the DANCE lunge classifier (R2=0.9, n=40). (H) Regression of the Divider assay lunge classifier score vs. manual score (R2=0.7739, n=40). (I) Precision, recall, and F1 scores of the DANCE lunge classifier compared with those of CADABRA and Divider.

Since aggressive lunges have a large dynamic range, we benchmarked our classifier across a range of aggressive behaviors and subdivided the ground-truth videos into four categories: 1. low aggressive, 0–70 lunges (Figure 2B); 2. moderately aggressive, 71–160 lunges (Figure 2C); 3. highly aggressive, 161–300 lunges (Figure 2D); and 4. hyperaggressive, >300 lunges (Figure 2E). DANCE scores remained comparable to ground-truth scores across all categories, whereas CADABRA and Divider underestimated the lunge counts (Figure 2B–E). Correlation analysis revealed a strong relationship between DANCE and ground-truth scores (Figure 2F, Supplementary File 2). In comparison, CADABRA and the Divider assay classifier showed a weaker correlation (Figure 2G-H, Supplementary File 2). We reasoned that CADABRA’s lower performance is most likely due to the rigid rules used to define a lunge (Dankert et al., 2009), whereas the Divider assay classifier, although also JAABA-based, was trained using data from a rectangular arena. Because JAABA classifiers rely on features influenced by the arena geometry, this mismatch likely reduced its accuracy in our circular setup. To further evaluate the performance, we computed the precision, recall, and F1 score (Figure 2I). The DANCE lunge classifier achieved a precision of 78.7%, recall of 73.1%, and an overall F1 score of 75.8%, exceeding the values obtained with other methods. Classifier robustness across multiple training videos, including the dataset used for inter-method comparisons (Video 9), is summarized in Figure 2—Figure Supplement 2.Together, our analysis suggests that the DANCE lunge classifier performs with high precision and quantifies lunge numbers robustly over a broad range of fighting intensities.

DANCE classifiers to quantify courtship behaviors in Drosophila

The first report of Drosophila courtship behavior described stereotypic behaviors such as wing “scissor-like” movements, with males “swaying around the female,” licking, tapping, and mounting (Sturtevant, 1915). By the 2000s, studies revealed genes and neural circuits involved in courtship (Dickson, 2008; Pavlou and Goodwin, 2013). Automated analysis techniques for courtship exist, but their adoption has been limited by expensive hardware, reliance on custom parts, or a lack of publicly available code and classifiers (Duistermars et al., 2018; Gil-Martí et al., 2023; Koemans et al., 2017; Reza et al., 2013; Tao et al., 2024).

MateBook is a recent rule-based pipeline for automating the quantification of courtship behavior (Ribeiro et al., 2018). It relies on predefined rules derived from CADABRA (Dankert et al., 2009). To resolve ambiguities in the two overlapping flies when their trajectories are estimated from the video recordings, identities are assigned by relying on the distinct body sizes of the male and female flies, as the females are larger than the males. This size assumption is problematic in assays that use decapitated virgin females, which approximate male size. Such conditions are often used to assess male courtship independent of female behavioral feedback, for example, when evaluating pheromone effects (Cook and Cook, 1975; Spieth, 1966). In these contexts, rule-based approaches can introduce false positives and false negatives because the body-size criterion is not met.

To overcome these limitations, we trained and validated five new courtship classifiers using JAABA (Kabra et al., 2013). These classifiers quantify distinct stages of the male courtship ritual (Sokolowski, 2001), including wing extension, following, circling, attempted copulation, and copulation. To ensure robustness across conditions, training datasets included videos with both decapitated and intact females (mated or virgin). We evaluated the DANCE classifier performance by comparing the outputs with the manual ground-truth and MateBook results. Because courtship behaviors differ in duration, we calculated a behavior index to enable direct comparisons between methods. Details of the classifier thresholds and criteria are provided in the Materials and Methods. To assess observer bias, annotations from two independent evaluators were compared, revealing no significant differences (Figure 2—Figure Supplement 1B–F).

Wing Extension

During unilateral wing extension, a male vibrates its wing at a specific frequency to produce a species-specific courtship song to attract the female (Shorey, 1962; Spieth, 1952).

Figure 3A shows wing extension indices derived from manual ground-truth (gray) and the DANCE wing extension classifier (orange) across 15 videos with decapitated virgin females; the two measures are comparable, whereas MateBook (purple) systematically reports lower scores in most videos (Figure 3A). The DANCE wing extension index is strongly correlated with the ground-truth scores (Figure 3B-C) but weakly correlated with MateBook (Figure 3B, D). The classifier performance metrics confirm high reliability (precision 92.2%, recall 98.1%, F1 95.1%) (Figure 3E). Similar trends were observed in the mated female dataset (Figure 3—Figure Supplement 1).

Evaluation of the DANCE wing extension classifier for quantifying courtship behavior.

(A) Wing extension index of males from 15 minutes long videos scored by manual ground-truth (gray), the DANCE wing extension classifier (orange) and MateBook (purple), with decapitated virgin females. MateBook underscored wing extension across multiple videos (Friedman’s ANOVA with Dunn’s test: ground-truth vs MateBook p=0.0020, n=15; ground-truth vs DANCE p>0.9999). (B) Comparison of ground-truth, DANCE, and MateBook wing extension scores (KruskallWallis ANOVA with Dunn’s test, ns, p>0.9999, *p=0.0436; n=15). (C) Regression analysis of the DANCE wing extension classifier vs. ground-truth (R2=0.9831, n=15). (D) Regression of MateBook vs. ground-truth (R2=0.1054, n=15). (E) Precision, recall, and F1 score of the DANCE wing extension classifier and MateBook relative ground-truth scores.

Attempted copulation and copulation

Copulation typically lasts for approximately 15–25 minutes, and its duration is primarily determined by the male (MacBean and Parsons, 1967). Interrupted mating experiments have shown that sperm are transferred several minutes after copulation begins (Fowler, 1973; Tompkins et al., 1980). This distinction separates mounting into two outcomes—successful copulation and unsuccessful attempted copulation—for which we developed separate classifiers (see Materials and Methods).

The DANCE attempted copulation classifier closely matched the ground-truth across videos (Figure 4A–B). Compared with MateBook, which often overestimates attempted copulation events, the DANCE classifier provides more consistent detection. The correlation with the ground-truth was stronger for DANCE as compared to MateBook (Figure 4C–D), and the performance metrics for DANCE were robust (precision 82.6%, recall 89.2%, F1 85.8%; Figure 4E).

Validation of the DANCE attempted-copulation classifier.

(A) Attempted copulation index of males from 15-minute-long videos scored by manual ground-truth (gray), the ‘DANCE attempted copulation classifier’ (orange) and the MateBook (purple) with both mated and decapitated females. (B) Comparison of ground-truth, DANCE attempted-copulation classifier, and MateBook scores (KruskallJWallis ANOVA with Dunn’s test, ns, p>0.9999; ****p<0.0001, n=32). (C) Regression analysis of the attempted-copulation classifier vs. ground-truth (R2=0.8742, n=32). (D) Regression analysis of MateBook vs. ground-truth (R2=1512, n=32). (E) Precision, recall, and F1 score of the DANCE and MateBook attempted-copulation classifiers relative to the ground-truth scores.

For copulation, which is trained on videos using mated and decapitated virgin females, the DANCE copulation classifier also matches the ground-truth with near-perfect performance (Figure 4—Figure Supplement 1). MateBook performs reasonably well for copulation, but alignment or arena-detection errors in some recordings cause occasional false negatives or positives (Figure 4—Figure Supplement 1A, videos #13 and #18).

Circling

A male circling around a female is a distinct courtship element implicated in female re-stimulation (Kessler, 1962). Variations in circling frequency across species contribute to reproductive isolation (Brown, 1965).

We evaluated the DANCE circling classifier on both the decapitated-virgin and mated-female datasets (Figure 5 and Figure 5—Figure Supplement 1). The circling indices from ground-truth and DANCE were comparable across videos (Figure 5A–B), whereas MateBook often under-represented circling in individual recordings (Figure 5A). The DANCE circling index correlated strongly with ground-truth scores as compared to MateBook (Figure 5C-D), and classifier metrics indicated robust performance (precision 98.0%, recall 92.1%, F1 95.0%; Figure 5E).

Evaluation of the DANCE circling classifier.

(A) Circling index of males from 15 minute long videos from the ground-truth (gray), ‘DANCE circling classifier’ (orange) and MateBook (purple) with decapitated virgin females Friedman’s ANOVA with Dunn’s test, p=0.2049, ****p<0.0001, n=12. (B) Comparison of the ground-truth, DANCE, and MateBook circling classifiers (ordinary one-way ANOVA with Dunnett’s test, ns, p=0.8014; *p=0.0157, n=12). (C) Regression analysis of the DANCE circling classifier vs. ground-truth (R2=0.92, n=12). (D) Regression of MateBook vs. ground-truth (R2=0.88, n=12). (E) Precision, recall, and F1 score of the DANCE and MateBook circling classifiers relative to the ground-truth score.

Following

During following, the male tracks the female’s movement to initiate subsequent courtship acts (Spieth, 1968). Because following is a relatively continuous and readily defined behavior, both the MateBook and the DANCE following classifier performed well (Figure 5—Figure Supplement 2A–D). However, the DANCE following classifier produced more balanced scores (precision 91.2%, recall 91.1%, F1 91.1%) compared with MateBook (precision 65.8%, recall 83.4%, F1 73.5%; Figure 5—Figure Supplement 2E), indicating lower rates of false positives and false negatives for DANCE.

Finally, we performed frame-level analyses in addition to bout-level evaluations to provide a more granular assessment of the courtship classifiers (see Materials and Methods). Frame-level metrics showed only marginal reductions in performance compared with bout-level metrics (Figure 5—Figure Supplement 3). Together, these results demonstrate that the DANCE classifiers provide a reliable and accurate means to quantify both aggression and courtship behaviors, supporting subsequent benchmarking using the DANCE hardware.

DANCE hardware

Existing setups for recording Drosophila aggression and courtship (Figure 1A; Figure 1—Figure Supplement 1–2) present several practical challenges that limit their broad adoption. These include the need for complex, custom-fabricated components, 3D-printed parts, specialized machine-vision cameras and backlights, and considerable technical expertise for data acquisition and processing (Chowdhury et al., 2021; Dankert et al., 2009; Gil-Martí et al., 2023; Koemans et al., 2017). Setting up some aggression assays (Dankert et al., 2009; Dierick, 2007) also requires coating chambers with fluon to prevent flies from walking on the walls, which is labor-intensive.

To provide a low-cost, easy-to-assemble alternative, we developed the DANCE hardware (Figure 6), an inexpensive, scalable, and robust system for recording Drosophila aggression and courtship behaviors.

DANCE hardware and recording setup.

(A) DANCE aggression setup (B) 3D-rendered components of the aggression setup (C) DANCE courtship setup (D) 3D-rendered components of the courtship setup, showing males and females separated by an X-ray film separator or ‘divider comb’. (E-G) Top and side views of the DANCE setup with a smartphone camera for recording and an electronic tablet as the backlight.

The DANCE hardware consists of readily available off-the-shelf components, including transparent medicine blister packs (tablet foils) used as recording chambers, which are mounted on 2 mm acrylic base plates and secured with paper tape (Figure 6A–D, Supplementary Videos 1–2). Instead of using machine-vision cameras, DANCE employs widely available Android smartphones for recording and substitutes backlights with tablets or smartphones displaying a white screen to provide uniform illumination (Figure 6E–G). Aggression and courtship behaviors were recorded at 30 fps and 1080p resolution.

For the aggression assays (Figure 6A–B, E), the blister foil was slid over a base plate containing an appleljuice agar food layer, which served as the interaction arena (Supplementary Videos 1–2). For courtship assays, the blister foil was bisected and fitted with a thin X-raylfilm separator comb that kept males and females apart until the start of recording, when the comb was removed to allow interaction (Supplementary Videos 3–4). Because the tablet or smartphone screens used as the backlight generate heat, we placed a transparent acrylic spacer above the backlight to create a 4 mm air gap for heat dissipation (Figure 6E–G; Supplementary Videos 2, 4). This modification was essential for maintaining consistent behavioral recordings (Figure 7).

Benchmarking DANCE hardware and application to neurogenetic tools.

(A-B) Courtship behaviors recorded using a pre-existing circular setup and DANCE setup in group-housed (GH) and single-housed (SH) flies for (C-D) wing extension, (C) ***p<0.0010, n=23; (D) ****p<0.0001, GH, n=22 and SH, n=33. (E-F) Attempted copulation, (E) ***p<0.0002, n=23; (F) **p<0.0022, GH, n=21 and SH, n=33. (G-H) Following, (G) ns, p>0.0959, n=23; (H) ns, p<0.2537, GH, n=22 and SH, n=32. (I-J) Circling, (I) *p<0.012, n=23; (J) *p<0.0104, GH, n=24 and SH, n=31. (K-L) Aggressive lunges were recorded using a pre-existing circular setup and a DANCE setup. (MlN) Lunges of SH flies compared with those of GH flies reared on food with yeast granules. (M) **p<0.0138, n=36; (N) **p<0.0372, n=40. (O) Effect of yeast extract food on aggressive behavior; ****p<0.0001, n=38. (P-Q) Genetic knockdown of the neuropeptide Drosulfakinin (Dsk) in insulin-producing neurons using dilp2-GAL4. (P) ns, p<0.0502; ns, p>0.9999; ****p<0.0001; **p<0.0040; n=35. (Q) ****p<0.0001, ns, p>0.9999, ****p<0.0001, *p>0.0210, n=30. (R) Optogenetic silencing of dopaminergic neurons with UAS-GtACR1 driven by the TH-GAL4 driver; ns, p<0.0986; ns, p>0.9999; ****p>0.0001; **p>0.0012; **p>0.0013; n=24. (ClJ and MlO) MannlWhitney U test; (PlR) KruskallWallis test with Dunn’s multiple comparisons.

The DANCE hardware is intentionally modular, allowing users to adapt the setup to their experimental needs. The detailed assembly, cleaning, and reuse protocols are provided in the Materials and Methods and on the GitHub page of the project.

Benchmarking DANCE hardware

We benchmarked the DANCE hardware by applying the validated DANCE classifiers to videos recorded in the DANCE setup and comparing the results with those from established recording systems. Wild-type males displayed quantitatively similar levels of courtship and aggression in DANCE arenas and in pre-existing setups (Figure 7; Supplementary Video 5).

To test whether DANCE reproduces established behavioral findings, we examined the effects of social isolation and enrichment on aggression and courtship. Previous studies have shown that single-housing (SH) increases courtship attempts (Dankert et al., 2009; Kim and Ehrman, 1998; Pan and Baker, 2014) and promotes aggression (Agrawal et al., 2020; Wang et al., 2008; Yadav et al., 2024). We found that both the DANCE and pre-existing setups captured similar and statistically significant differences in courtship behaviors between SH and group-housed (GH) males (Figure 7C–J). These results confirm that DANCE reliably detects behavioral modulation by social experience.

We next compared an established aggression assay (Dankert et al., 2009); Figure 7K) with the DANCE aggression setup (Figure 7L). Both systems detected aggressive lunges in SH flies and showed consistent differences between SH and GH conditions (Figure 7M–N). Thus, DANCE hardware provides comparable sensitivity to conventional machine-vision-based setups while being more accessible.

We also tested whether diet composition alters aggression in DANCE assays, as nutrient availability and microbiome interactions can influence male aggression (Jia et al., 2021; Lim et al., 2014). Replacing yeast granules in the diet with yeast extract powder reduced baseline aggression (Figure 7O). These findings demonstrate that the DANCE hardware is sensitive enough to detect diet-dependent behavioral differences.

To test whether DANCE is compatible with neurogenetic manipulations, we used RNAi-mediated knockdown of the neuropeptide Drosulfakinin (Dsk) in insulin-producing neurons using the dilp2-GAL4 driver. Consistent with our previous findings (Agrawal et al., 2020), SH males with Dsk knockdown exhibited significantly increased aggressive lunges compared to controls (Figure 7P–Q).

We then evaluated DANCE’s suitability for optogenetic assays, a common approach to dissect neural circuits underlying aggression (Hoopfer et al., 2015; Wohl et al., 2023; Yadav et al., 2024). An earlier study has shown that constitutive silencing of broad populations of dopaminergic neurons using TH-GAL4 produces unhealthy flies with impaired locomotion that rarely fight (Alekseyenko et al., 2013). We reasoned that transient, light-controlled silencing could overcome these limitations and tested this using an optogenetic module integrated with DANCE (Figure 7—Figure Supplement 1A–B; DANCE GitHub).

Optogenetic silencing of dopaminergic neurons expressing the green-light-sensitive anion channelrhodopsin GtACR1 (Govorunova et al., 2015; Mohammad et al., 2017) during 20-minute interactions resulted in a significant increase in aggressive lunges in SH flies (Figure 7R). In addition, we observed higher frequencies of wing flicks and high-intensity aggressive behaviors such as boxing and tussling (Supplementary Video 6). Importantly, continuous silencing for 12 hours did not alter general locomotor activity between the GH and SH flies (Figure 7—Figure Supplement 2), confirming that these effects were not due to impaired movement.

We also examined whether arena size influenced behavior. TH-GAL4 > UAS-GtACR1 males showed increased lunging versus controls across all sizes tested (13, 17 and 21 mm; Figure 7—Figure Supplement 3A–C), indicating the optogenetic effect was robust to chamber dimensions. In courtship assays (11, 13 and 17 mm arenas; Figure 7—Figure Supplement 4A–L), single-housed males exhibited more wing extension, attempted copulation and circling than group-housed males, while following was similar. Although statistical tests were limited to within-size comparisons, the distributions suggest a modest decrease in interaction frequency in larger arenas, consistent with previous observations that larger chambers reduce encounter rates (Chowdhury et al., 2021).

Taken together, these results show that DANCE hardware provides reliable, reproducible behavioral measurements and is compatible with genetic and optogenetic manipulations as well as environmental perturbations.

Discussion

Here, we present the DANCE assay, an easy-to-use, modular, and robust analysis pipeline with inexpensive hardware to record and quantify aggression and courtship behaviors. We developed six novel behavioral classifiers using supervised machine learning to accurately quantify the aggression and courtship behaviors of Drosophila males. The hardware component of DANCE, fabricated from repurposed and low-cost materials, provides a practical alternative to conventional machine-vision setups. The DANCE setup can be built for less than 0.30 USD, representing an approximately 10,000-fold reduction in cost compared with standard systems. Despite this simplicity, its performance was comparable to that of more specialized and expensive hardware, validating DANCE as a reliable and accessible behavioral platform. This accessibility enables rapid behavioral screening and wider adoption by the neuroscience community, including resource-limited laboratories and teaching environments.

Various components of the DANCE assay, such as the behavioral classifiers, hardware design, and analysis code, are publicly available and can be used independently. This open, modular design gives researchers the flexibility to customize classifiers for specific behavioral paradigms or incorporate new data without developing a classifier from scratch. Although not implemented here, DANCE can be readily extended to real-time feedback experiments using open-source reactive programming tools such as Bonsai (Lopes et al., 2015). This framework can also support the development of future classifiers for additional social behaviors, including aggressive acts such as fencing, wing flicking, tussling, chasing, or female headbutting (Chen et al., 2002; Nilsen et al., 2004), and courtship behaviors such as male tapping, licking, or female rejection (Sokolowski, 2001).

Such high-resolution analysis of complex social interactions can provide deeper understanding of mating dynamics, sexual selection, and the influence of genetics and evolution. Further, quantifying behavioral dynamics can reveal the temporal organization of individual components of behavior (Nilsen et al., 2004; Seeds et al., 2014; Simon and Heberlein, 2020; Zhang et al., 2020). DANCE can also serve as a flexible framework for studying complex behaviors across multiple Drosophila species and other insects, enabling comparative and evolutionary analyses. The adaptability and portability of the DANCE assay make it particularly useful for ethologists examining insect behavior in semi-natural or field-like environments. Together, these features position DANCE as a bridge between laboratory-based ethology and ecological studies of natural behavior.

Materials and Methods

The details of all the custom codes, analysis pipelines, sample files used to run the analysis and DANCE classifiers are available in our GitHub repository at https://github.com/agrawallab/DANCE.

Fly Husbandry

Flies were reared on standard food at 25°C and 65% relative humidity with a 12 hr:12 hr lightldark cycle. All the assays were performed at 25°C with 65% relative humidity, unless mentioned otherwise. For the aggression and courtship experiments, Canton-S male flies were collected within 24 hours of eclosion and housed in groups (20 male flies per vial, 90 mm in length and 25 mm in diameter) or isolated (1 male fly per vial, 70 mm in length and 10 mm in diameter) for 6 days.

The following fly lines were acquired from the Bloomington Drosophila Stock Center (BDSC), USA: TH-GAL4 (RRID:BDSC_51982), Dilp2-GAL4 (RRID:BDSC_37516), Dsk-RNAi (RRID:BDSC_25869) and attP2 empty vector control (RRID:BDSC_36303). UAS-GtACR1 flies were a gift from Gaurav Das, NCCS, Pune, India, and Canton-S (CS) flies were obtained from Ulrike Heberlein, HHMI, Janelia Research Campus, USA.

Aggression assay

Aggression assays were performed as described previously (Dankert et al., 2009; Dierick, 2007). In brief, the behavioral chamber is made up of 12 well-aggressive arenas (10 mm in height and 16 mm in diameter/arena). These arenas were covered by a sliding lid with 2 mm loading holes to facilitate the introduction of flies. A pair of male flies that were housed either as a group (GH) or single fly (SH) were introduced into the arena wells by gentle mouth aspiration through the loading holes. After the flies were loaded, the sliding lid was tightened with screws. Fluon (Insect-a-slip, Bioquip: cat #2871B) was applied to the arena walls, which were left to dry overnight to create a slippery surface and prevent climbing. Sigmacote (SigmalAldrich: SL2) was used to coat the sliding lid to reduce walking on the arena ceiling. The chamber was placed on a food plate containing commercial apple juice (without added sugars), 2.5% w/v sucrose (HiMedia: GRM601), and 2.25% w/v agar (HiMedia: GRM026). For experiments to test the effects of fly food nutrients, either 2.4% yeast extract powder (HiMedia: RM0271) or 2.4% yeast granules (Prime Instant Dry yeast, AB Maury, India) were mixed in the fly food. For optogenetic experiments, LEDs emitting 520–540 nm light (Lumileds: 2835) were used at an intensity of 0.0004 µW, which was measured with a power meter (Newport: 843-R). For all the aggression assays, the flies were allowed to acclimatize in the arena for 5 minutes, after which the activity was recorded for 20 minutes. The assays were performed during ZT0 - ZT2.5, i.e., during the first 2.5 hours of the morning activity peak.

Courtship assay

Courtship assays were performed as described previously (Koemans et al., 2017; Ribeiro et al., 2018). Single pairs of males and females were introduced into individual arenas of an 18-well courtship chamber (10 mm diameter). Male and female flies were introduced into one half of the chamber by sliding entry holes with a removable separator that divided the chamber into two halves. The flies were allowed to acclimatize to the arena for 5 minutes, after which the separator was removed, and courtship behavior was recorded for 15 minutes at 30 fps using a white backlight. The assay was performed from ZT0 to ZT3 or ZT9 to ZT12 (during peak activity windows). For mated females, 20 females were housed with 10 males for 4–6 days. For decapitated virgin females, 2–4-day-old virgins were anesthetized with CO₂ and decapitated immediately before the assay.

DANCE hardware

Circular transparent medicine blister packs serve as aggression or courtship arenas. Blister packs were mounted on 2 mm acrylic base plates and secured with paper tape. The arena dimensions were as follows: aggression, 13 mm × 5.5 mm (diameter × height) (also tested: 17 × 4 mm and 21 × 8 mm); courtship, 11 mm × 4.5 mm (also tested: 13 × 5.5 mm and 17 × 4 mm). Arena walls and roofs were coated with Sigmacote (SigmalAldrich: SL2) using cotton swabs (Solimo, Amazon India) to prevent flies from climbing.

A thin 2 mm acrylic base plate carrying the food layer (2.25% w/v agar in commercial apple juice with 2.5% w/v sucrose) and a side spacer were assembled and held together with paper tape (Supplementary Video 1). The tip of the food plate was covered with paper tape to allow smooth sliding of the blister foil without damaging the food surface. Flies were introduced through loading holes (2 mm diameter), and paper tape strips were used to prevent food damage during assembly and to seal small gaps to avoid escape (Supplementary Video 2).

Arenas were reusable for approximately 30–40 times when washed in 0.05% Alcojet (Alconox: 1401-1) and air-dried. Heat exposure during cleaning was avoided to prevent deformation.

The standard DANCE courtship arena used transparent medicine blister packs measuring 11 mm in diameter and 4.5 mm in height. Additional arena sizes of 13 × 5.5 mm and 17 × 4 mm (diameter × height) were also tested. The top surface of each arena strip (five wells in a row) was modified by cutting thin slits with a sharp razor to insert a separator comb made from repurposed X-ray film. The foil and 2 mm acrylic base plate, which contained a loading hole (2 mm diameter), were joined with paper tape. The courtship base plate contained two loading holes so that male and female flies could be introduced on either side of the separator comb. The assembly was sealed at both ends with paper tape to prevent gaps from escaping (Supplementary Video 3).

Before recording, the separator comb was gently lifted to allow interaction between the flies, taking care to keep the slit edges intact and avoid distortion during filming (Supplementary Video 4). Blister packs modified with slits and separator combs are typically reusable for ∼10–15 experiments, after which the slits tend to widen, and new blister packs are recommended.

Video acquisition

For the traditional setup, the interaction of the flies was recorded using machine-vision cameras (DMK 33UX252 USB 3.0 monochrome camera). White backlight (TMS, BHS4-00-100-X-W-24V) provided the light source for both the courtship and aggression experiments. Videos were recorded at 30 frames per second (fps) for 15 minutes for courtship or 20 minutes for aggression in H.264 (.mp4) format with 1440×1080 resolution. These videos were used for training, testing, and validating the DANCE classifiers.

For DANCE hardware testing, various Android smartphone cameras were used (Huawei Y9 2019; OnePlus Nord CE 2 Lite 5G, model: CPH2381; Redmi Note 11 Pro+ 5G, model: 221116SI) at 30 fps, 1080p resolution in H.264 (.mp4) format for 15 minutes (courtship) or 20 minutes (aggression). An electronic tablet (iPad Air, 5th Generation) or smartphone (iPhone 13) running a ‘white screen light app’ served as the background illumination source. The screen brightness was adjusted within the app to optimize contrast for fly tracking. A transparent acrylic sheet with 4 mm spacers was kept on top of this ‘backlight’ to create an air gap to ensure heat exchange and prevent the DANCE arenas from becoming hot. Devices were recommended to be placed in airplane mode during recordings to avoid interruptions.

Tracking flies using FlyTracker

Fly locations, body orientations, and interactions were tracked using Caltech FlyTracker (Eyjolfsdottir et al., 2014). These data were then pushed to the JAABA pipeline to develop DANCE classifiers. Occurrences of identity switches were corrected using the FlyTracker “visualizer” identity-correction tool. Tracking accuracy and identity swap quantification were validated by a semi-manual inspection of flagged frames and intervals (see Supplementary File 5 for details).

Pre-existing algorithms used for benchmarking

CADABRA

CADABRA (Dankert et al., 2009) analyzes two-fly interactions based on spatial and postural features, classifying lunges, wing threats, circling, wing extension, and copulation based on fixed rules. The specific CADABRA definitions are described in the subsequent section.

Divider Assay

The Divider Assay (Chowdhury et al., 2021) uses a 3D-printed rectangular chamber with 12 arenas (13lmm × 4.5lmm, W × H), each separated by an opaque divider. Behavioral data are analyzed with a custom FlyTracker-JAABA pipeline. Since the Divider assay classifier was trained on recordings from a rectangular geometry, it can affect transferability to circular arenas.

MateBook

MateBook (Ribeiro et al., 2018); https://github.com/Dicksonlab/MateBook) uses machine vision to track flies and classify male courtship behaviors (following, wing extension, orientation, copulation, and circling). The outputs include a .tsv file with bout statistics and an ethogram. For comparison with the DANCE classifiers, the MateBook persistence filters were adjusted so that the minimum bout duration threshold was 0.33 s (10 frames at 30 fps) for all behaviors except copulation, which retained a 45 s threshold. This adjustment ensured comparable persistence criteria between the MateBook and DANCE analyses.

Developing DANCE classifiers

JAABA (Kabra et al., 2013) was used to train classifiers iteratively: true bouts were labeled, obvious non-bouts assigned as “None,” and false positives were relabeled until performance plateaued.

The DANCE lunge classifier was trained on 11 independent videos with classifier accuracy improving progressively from Video 1 to Video 9. Peak performance was achieved after inclusion of the ninth training video, which provided the best balance of precision, recall, and F1 score. Adding further data (Videos 10 and 11) did not enhance classifier accuracy and instead produced a slight reduction in precision and recall, likely due to increased behavioral variability across sessions. Therefore, the final lunge classifier was trained on nine videos, which yielded the most robust and generalizable model.

Courtship classifiers were developed following the same iterative procedure. Training sets included videos with both decapitated and intact females (mated or virgin) to capture behavioral variability and ensure robust generalization. Independent classifiers were trained for wing extension, following, circling, attempted copulation, and copulation. All classifiers were validated using manually annotated “ground-truth” test videos that were not included in the training set.

Courtship training sets included decapitated virgin videos to enrich attempted copulation and circling bouts and to balance “None” class examples for following and copulation. Wing extension used mated females only.

The training data volumes (frames and approximate durations) were as follows:

Test videos were manually annotated in JAABA ground-truthing mode before any classifier predictions were examined, and annotations were performed independently of classifier outputs.

Manual behavioral annotations and inter-annotator reliability

Manual annotations (“ground-truth”) were generated using JAABA’s ground-truthing mode by labeling behavioral bouts frame-by-frame. To assess observer bias, subsets of videos were annotated independently by two evaluators and compared using non-parametric tests. Where relevant, the results describe inter-annotator comparisons.

Characterization of male aggression and courtship behaviors

The DANCE classifiers were trained and validated in JAABA using established behavioral definitions from previous studies (Dankert et al., 2009; Ribeiro et al., 2018), with input from experienced users.

Behavior definitions

Lunge

As defined by (Dankert et al., 2009; Ribeiro et al., 2018), “the attacking fly rises on hind legs, lifting its long body axis by 45°, then snaps down on its opponent’s body with its head at ∼200 mm/s.”

Wing extension

The angle between the body axis and wing tip line exceeds 30°, persisting for at least 13 frames (0.5 s at 25 fps) (Dankert et al., 2009; Ribeiro et al., 2018).

Attempted copulation

Abdominal curling without mounting, or mounting lasting ≥0.33 s but <45 s (10–1350 frames at 30 fps).

Copulation

Mounting lasting ≥45 s (>1350 frames at 30 fps), adapted from (Dankert et al., 2009; Ribeiro et al., 2018).

Circling

Sideways drift around the female in a circular path at constant velocity, persisting for ≥13 frames (Dankert et al., 2009; Ribeiro et al., 2018).

Following

The male remains 2–5 mm behind the female while both walk at ≥2 mm/s, persisting for ≥25 frames (Dankert et al., 2009; Ribeiro et al., 2018).

To account for variability in female mating condition and body size, training datasets included videos of males paired with decapitated virgin, intact virgin, and mated females. For the copulation classifier, videos in which males were paired with virgin females to capture prolonged occlusion events characteristic of mating were used for training.

For duration-based classifiers, a post-processing filter was applied to exclude bouts shorter than 98% of those observed in manual annotations, ensuring consistency with human-defined behavioral durations.

This framework can be readily adapted by the research community to develop additional behavioral classifiers. Owing to file size limitations, training videos are not hosted online but are available upon request.

Manual behavioral annotations

To quantitatively evaluate classifier performance, manual behavioral annotations (“ground-truth”) were generated using JAABA’s ground-truthing mode (Kabra et al., 2013). Classifier robustness was assessed on unseen videos comprising the testing set. Each testing video was first manually annotated by identifying behavioral bouts as “true behavior,” independent of the classifier’s output. These same videos were then processed through the trained classifier using JAABAPlot, and the results were compared as described below.

Comparison of manual and DANCE annotations

The classifier outputs were compared with manually annotated ground-truth data at both the bout and frame levels, depending on the classifier type. For the single-frame lunge classifier, comparisons were based on total bout counts, whereas for all duration-based courtship classifiers (wing extension, following, circling, attempted copulation, and copulation), comparisons were made using bout durations and frame level annotations.

For each assay, a behavioral index was calculated as the proportion of frames in which the male engaged in the specified behavior. This was obtained by dividing the total number of frames annotated for that behavior by the total number of frames in the recording. Regression analyses and performance metrics were computed using either bout counts or behavioral indices, depending on the classifier type, in GraphPad Prism 8 (GraphPad Software). Manual annotations and classifier outputs were compared to identify true positives (TP), false positives (FP), and false negatives (FN), at either frame-level or bout-level.

Bout-level analysis

A predicted bout was scored as a TP if it overlapped with a ground-truth bout by at least one frame (∼33 ms at 30 fps), consistent with previous studies (Leng et al., 2020). When multiple predicted bouts overlapped a single ground-truth bout, they were collectively counted as one TP. Conversely, when a single predicted bout overlapped multiple ground-truth bouts, the TP count equaled the number of ground-truth bouts. Predicted bouts with no overlap were scored as FP, and ground-truth bouts with no overlapping prediction were scored as FN. These same criteria were applied to both bout-level and frame-level evaluations, with the latter accounting for the total number of frames contributing to TP, FP, and FN classifications to provide a more granular measure of accuracy (see Figure 5—Figure Supplement 3).

Frame-level analysis

A predicted frame was scored TP if it matched the ground-truth frame; frames predicted as behavior only by the classifier were FP; frames annotated as behavior only by ground-truth were FN. These frame-level TP, FP, FN were then similarly used to calculate precision, recall, and F1 score, providing a more granular measure of classifier accuracy.

Performance metrics

A custom Python script was used to calculate overlaps and derive standard classification metrics. The precision, recall, and F1 scores were computed using the following formulas: Precision represents the fraction of correctly predicted positive observations among all predicted positives, recall represents the fraction of correctly predicted positive observations among all actual positives, and the F1 score provides the harmonic mean of precision and recall. These metrics were used to quantify and compare the performance of the DANCE classifiers against manually annotated ground-truth and existing algorithms.

For duration-based behaviors, the behavioral index was used as a continuous variable in regression and performance analyses. For the lunge classifier, which identifies discrete one-frame events, comparisons were made using total bout counts.

Statistical analysis

All the statistical analyses were performed using GraphPad Prism 8, custom Python scripts, or Microsoft Excel. For non-normally distributed data, non-parametric tests such as the Mann–Whitney U test or Kruskal–Wallis ANOVA with appropriate post hoc corrections were used.